% language=us runpath=texruns:manuals/luametatex

\environment luametatex-style

\startcomponent luametatex-libraries

\startchapter[reference=libraries,title={Extra libraries}]

\startsection[title=Introduction]

The libraries can be grouped in categories like fonts, languages, \TEX,
\METAPOST, \PDF, etc. There are however also some that are more general purpose
and these are discussed here.

\stopsection

\startsection[title=File and string readers: \type {fio} and type {sio}]

This library provides a set of functions for reading numbers from a file and
in addition to the regular \type {io} library functions. The following
work on normal \LUA\ file handles.

\starttabulate[|Tw(12em)|T|T|]
\DB name        \BC arguments \BC results \NC \NR
\TB
\NC readcardinal1     \NC (f)      \NC a 1 byte unsigned integer \NC \NR
\NC readcardinal2     \NC (f)      \NC a 2 byte unsigned integer \NC \NR
\NC readcardinal3     \NC (f)      \NC a 3 byte unsigned integer \NC \NR
\NC readcardinal4     \NC (f)      \NC a 4 byte unsigned integer \NC \NR
\NC readcardinaltable \NC (f,n,b)  \NC \type {n} cardinals of \type {b} bytes \NC \NR
\NC readinteger1      \NC (f)      \NC a 1 byte signed integer \NC \NR
\NC readinteger2      \NC (f)      \NC a 2 byte signed integer \NC \NR
\NC readinteger3      \NC (f)      \NC a 3 byte signed integer \NC \NR
\NC readinteger4      \NC (f)      \NC a 4 byte signed integer \NC \NR
\NC readintegertable  \NC (f,n,b)  \NC \type {n} integers of \type {b} bytes \NC \NR
\NC readfixed2        \NC (f)      \NC a float made from a 2 byte fixed format \NC \NR
\NC readfixed4        \NC (f)      \NC a float made from a 4 byte fixed format \NC \NR
\NC read2dot14        \NC (f)      \NC a float made from a 2 byte in 2dot4 format \NC \NR
\NC setposition       \NC (f,p)    \NC goto position \type {p} \NC \NR
\NC getposition       \NC (f)      \NC get the current position \NC \NR
\NC skipposition      \NC (f,n)    \NC skip \type {n} positions \NC \NR
\NC readbytes         \NC (f,n)    \NC \type {n} bytes \NC \NR
\NC readbytetable     \NC (f,n)    \NC \type {n} bytes\NC \NR
\LL
\stoptabulate

When relevant there are also variants that end with \type {le} that do it the
little endian way. The fixed and dot floating points formats are found in font
files and return \LUA\ doubles.

A similar set of function as in the \type {fio} library is available in the \type
{sio} library: \libidx {sio} {readcardinal1}, \libidx {sio} {readcardinal2},
\libidx {sio} {readcardinal3}, \libidx {sio} {readcardinal4}, \libidx {sio}
{readcardinaltable}, \libidx {sio} {readinteger1}, \libidx {sio} {readinteger2},
\libidx {sio} {readinteger3}, \libidx {sio} {readinteger4}, \libidx {sio}
{readintegertable}, \libidx {sio} {readfixed2}, \libidx {sio} {readfixed4},
\libidx {sio} {read2dot14}, \libidx {sio} {setposition}, \libidx {sio}
{getposition}, \libidx {sio} {skipposition}, \libidx {sio} {readbytes} and
\libidx {sio} {readbytetable}. Here the first argument is a string instead of a
file handle.

\stopsection

\startsection[title=\type{md5}]

\starttabulate[|Tw(12em)|T|T|]
\DB name \BC arguments \BC results \NC \NR
\TB
\NC sum  \NC \NC \NC \NR
\NC hex  \NC \NC \NC \NR
\NC HEX  \NC \NC \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title=\type{sha2}]

\starttabulate[|Tw(12em)|T|T|]
\DB name      \BC arguments \BC results \NC \NR
\TB
\NC digest256 \NC \NC \NC \NR
\NC digest384 \NC \NC \NC \NR
\NC digest512 \NC \NC \NC \NR
\LL
\stoptabulate

\stopsection

% \startsection[title=\type{flate}]
%
% \starttabulate[|Tw(12em)|T|T|]
% \DB name             \BC arguments \BC results \NC \NR
% \TB
% \NC flate_compress   \NC \NC \NC \NR
% \NC flate_decompress \NC \NC \NC \NR
% \NC zip_compress     \NC \NC \NC \NR
% \NC zip_decompress   \NC \NC \NC \NR
% \NC gz_compress      \NC \NC \NC \NR
% \NC gz_decompress    \NC \NC \NC \NR
% \NC update_adler32   \NC \NC \NC \NR
% \NC update_crc32     \NC \NC \NC \NR
% \LL
% \stoptabulate
%
% \stopsection

\startsection[title=\type{xzip}]

\starttabulate[|Tw(12em)|T|T|]
\DB name             \BC arguments \BC results \NC \NR
\TB
\NC compress   \NC \NC \NC \NR
\NC decompress \NC \NC \NC \NR
\NC adler32    \NC \NC \NC \NR
\NC crc32      \NC \NC \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title=\type{xmath}]

This library just opens up standard \CCODE\ math library and the main reason for
it being there is that it permits advanced graphics in \METAPOST\ (via the \LUA\
interface). There are three constant values:

\starttabulate[|Tw(12em)|T|T|]
\DB name \BC arguments \BC results \NC \NR
\TB
\NC inf  \NC \emdash \NC \cldcontext{xmath.inf} \NC \NR
\NC nan  \NC \emdash \NC \cldcontext{xmath.nan} \NC \NR
\NC pi   \NC \emdash \NC \cldcontext{xmath.pi}  \NC \NR
\LL
\stoptabulate

and a lot of functions:

\starttabulate[|Tw(12em)|T|T|]
\DB name       \BC arguments \BC results \NC \NR
\TB
\NC acos       \NC (a)       \NC \NC \NR
\NC acosh      \NC (a)       \NC \NC \NR
\NC asin       \NC (a)       \NC \NC \NR
\NC asinh      \NC (a)       \NC \NC \NR
\NC atan       \NC (a[,b])   \NC \NC \NR
\NC atan2      \NC (a[,b])   \NC \NC \NR
\NC atanh      \NC (a)       \NC \NC \NR
\NC cbrt       \NC (a)       \NC \NC \NR
\NC ceil       \NC (a)       \NC \NC \NR
\NC copysign   \NC (a,b)     \NC \NC \NR
\NC cos        \NC (a)       \NC \NC \NR
\NC cosh       \NC (a)       \NC \NC \NR
\NC deg        \NC (a)       \NC \NC \NR
\NC erf        \NC (a)       \NC \NC \NR
\NC erfc       \NC (a)       \NC \NC \NR
\NC exp        \NC (a)       \NC \NC \NR
\NC exp2       \NC (a)       \NC \NC \NR
\NC expm1      \NC (a)       \NC \NC \NR
\NC fabs       \NC (a)       \NC \NC \NR
\NC fdim       \NC (a,b)     \NC \NC \NR
\NC floor      \NC (a)       \NC \NC \NR
\NC fma        \NC (a,b,c)   \NC \NC \NR
\NC fmax       \NC (...)     \NC \NC \NR
\NC fmin       \NC (...)     \NC \NC \NR
\NC fmod       \NC (a,b)     \NC \NC \NR
\NC frexp      \NC (a,b)     \NC \NC \NR
\NC gamma      \NC (a)       \NC \NC \NR
\NC hypot      \NC (a,b)     \NC \NC \NR
\NC isfinite   \NC (a)       \NC \NC \NR
\NC isinf      \NC (a)       \NC \NC \NR
\NC isnan      \NC (a)       \NC \NC \NR
\NC isnormal   \NC (a)       \NC \NC \NR
\NC j0         \NC (a)       \NC \NC \NR
\NC j1         \NC (a)       \NC \NC \NR
\NC jn         \NC (a,b)     \NC \NC \NR
\NC ldexp      \NC (a,b)     \NC \NC \NR
\NC lgamma     \NC (a)       \NC \NC \NR
\NC l0         \NC (a)       \NC \NC \NR
\NC l1         \NC (a)       \NC \NC \NR
\NC ln         \NC (a,b)     \NC \NC \NR
\NC log        \NC (a[,b])   \NC \NC \NR
\NC log10      \NC (a)       \NC \NC \NR
\NC log1p      \NC (a)       \NC \NC \NR
\NC log2       \NC (a)       \NC \NC \NR
\NC logb       \NC (a)       \NC \NC \NR
\NC modf       \NC (a,b)     \NC \NC \NR
\NC nearbyint  \NC (a)       \NC \NC \NR
\NC nextafter  \NC (a,b)     \NC \NC \NR
\NC pow        \NC (a,b)     \NC \NC \NR
\NC rad        \NC (a)       \NC \NC \NR
\NC remainder  \NC (a,b)     \NC \NC \NR
\NC remquo     \NC (a,b)     \NC \NC \NR
\NC round      \NC (a)       \NC \NC \NR
\NC scalbn     \NC (a,b)     \NC \NC \NR
\NC sin        \NC (a)       \NC \NC \NR
\NC sinh       \NC (a)       \NC \NC \NR
\NC sqrt       \NC (a)       \NC \NC \NR
\NC tan        \NC (a)       \NC \NC \NR
\NC tanh       \NC (a)       \NC \NC \NR
\NC tgamma     \NC (a)       \NC \NC \NR
\NC trunc      \NC (a)       \NC \NC \NR
\NC y0         \NC (a)       \NC \NC \NR
\NC y1         \NC (a)       \NC \NC \NR
\NC yn         \NC (a)       \NC \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title=\type{xcomplex}]

\LUAMETATEX\ also provides a complex library \type {xcomplex}. The complex
number is a userdatum:

\starttabulate[|Tw(12em)|T|T|]
\DB name       \BC arguments \BC results \NC \NR
\TB
\NC new        \NC (r,i)     \NC a complex userdata type \NC \NR
\NC tostring   \NC (z)       \NC a string representation \NC \NR
\NC topair     \NC (z)       \NC two numbers \NC \NR
\LL
\stoptabulate

There is a bunch of functions that take a complex number:

\starttabulate[|Tw(12em)|T|T|]
\DB name       \BC arguments \BC results \NC \NR
\TB
\NC abs       \NC (a)       \NC \NC \NR
\NC arg       \NC (a)       \NC \NC \NR
\NC imag      \NC (a)       \NC \NC \NR
\NC real      \NC (a)       \NC \NC \NR
\NC onj       \NC (a)       \NC \NC \NR
\NC proj      \NC (a)       \NC \NC \NR
\NC exp"      \NC (a)       \NC \NC \NR
\NC log       \NC (a)       \NC \NC \NR
\NC sqrt      \NC (a)       \NC \NC \NR
\NC pow       \NC (a,b)     \NC \NC \NR
\NC sin       \NC (a)       \NC \NC \NR
\NC cos       \NC (a)       \NC \NC \NR
\NC tan       \NC (a)       \NC \NC \NR
\NC asin      \NC (a)       \NC \NC \NR
\NC acos      \NC (a)       \NC \NC \NR
\NC atan      \NC (a)       \NC \NC \NR
\NC sinh      \NC (a)       \NC \NC \NR
\NC cosh      \NC (a)       \NC \NC \NR
\NC tanh      \NC (a)       \NC \NC \NR
\NC asinh     \NC (a)       \NC \NC \NR
\NC acosh     \NC (a)       \NC \NC \NR
\NC atanh     \NC (a)       \NC \NC \NR
\LL
\stoptabulate

These are accompanied by \type {libcerf} functions:

\starttabulate[|Tw(12em)|T|Tpl|]
\DB name       \BC arguments \BC results \NC \NR
\TB
\NC erf        \NC (a)       \NC The complex error function erf(z) \NC \NR
\NC erfc       \NC (a)       \NC The complex complementary error function erfc(z) = 1 - erf(z) \NC \NR
\NC erfcx      \NC (a)       \NC The underflow-compensating function erfcx(z) = exp(z^2) erfc(z) \NC \NR
\NC erfi       \NC (a)       \NC The imaginary error function erfi(z) = -i erf(iz) \NC \NR
\NC dawson     \NC (a)       \NC Dawson's integral D(z) = sqrt(pi)/2 * exp(-z^2) * erfi(z) \NC \NR
\NC voigt      \NC (a,b,c)   \NC The convolution of a Gaussian and a Lorentzian \NC \NR
\NC voigt_hwhm \NC (a,b)     \NC The half width at half maximum of the Voigt profile \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title=\type{xdecimal}]

As an experiment \LUAMETATEX\ provides an interface to the \type {decNumber}
library that we have on board for \METAPOST\ anyway. Apart from the usual
support for operators there are some functions.

\starttabulate[|Tw(12em)|T|T|]
\DB name         \BC arguments  \BC results \NC \NR
\TB
\NC abs          \NC (a)        \NC \NC \NR
\NC new          \NC ([n or s]) \NC \NC \NR
\NC copy         \NC (a)        \NC \NC \NR
\NC trim         \NC (a)        \NC \NC \NR
\NC tostring     \NC (a)        \NC \NC \NR
\NC tonumber     \NC (a)        \NC \NC \NR
\NC setprecision \NC (n)        \NC \NC \NR
\NC getprecision \NC ()         \NC \NC \NR
\NC conj         \NC (a)        \NC \NC \NR
\NC abs          \NC (a)        \NC \NC \NR
\NC pow          \NC (a,b)      \NC \NC \NR
\NC sqrt         \NC (a)        \NC \NC \NR
\NC ln           \NC (a)        \NC \NC \NR
\NC log          \NC (a)        \NC \NC \NR
\NC exp          \NC (a)        \NC \NC \NR
\NC bor          \NC (a,b)      \NC \NC \NR
\NC bxor         \NC (a,b)      \NC \NC \NR
\NC band         \NC (a,b)      \NC \NC \NR
\NC shift        \NC (a,b)      \NC \NC \NR
\NC rotate       \NC (a,b)      \NC \NC \NR
\NC minus        \NC (a)        \NC \NC \NR
\NC plus         \NC (a)        \NC \NC \NR
\NC min          \NC (a,b)      \NC \NC \NR
\NC max          \NC (a,b)      \NC \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title=\type{lfs}]

The original \type {lfs} module has been adapted a bit to our needs but for
practical reasons we kept the namespace. This module will probably evolve a bit
over time.

\starttabulate[|Tw(12em)|T|Tp|]
\DB name              \BC arguments \BC results \NC \NR
\TB
\NC attributes        \NC (name) \NC \NC \NR
\NC chdir             \NC (name) \NC \NC \NR
\NC currentdir        \NC ()     \NC \NC \NR
\NC dir               \NC (name) \NC \type {name}, \type {mode}, \type {size} and \type {mtime} \NC \NR
\NC mkdir             \NC (name) \NC \NC \NR
\NC rmdir             \NC (name) \NC \NC \NR
\NC touch             \NC (name) \NC \NC \NR
\NC link              \NC (name) \NC \NC \NR
\NC symlinkattributes \NC (name) \NC \NC \NR
\NC isdir             \NC (name) \NC \NC \NR
\NC isfile            \NC (name) \NC \NC \NR
\NC iswriteabledir    \NC (name) \NC \NC \NR
\NC iswriteablefile   \NC (name) \NC \NC \NR
\NC isreadabledir     \NC (name) \NC \NC \NR
\NC isreadablefile    \NC (name) \NC \NC \NR
\LL
\stoptabulate

The \type {dir} function is a traverser which in addition to the name returns
some more properties. Keep in mind that the traverser loops over a directory and
that it doesn't run well when used nested. This is a side effect of the operating
system. It is also the reason why we return some properties because querying them
via \type {attributes} would interfere badly.

The following attributes are returned by \type {attributes}:

\starttabulate[|Tw(12em)|T|]
\DB name         \BC value \NC \NR
\TB
\NC mode         \NC \NC \NR
\NC size         \NC \NC \NR
\NC modification \NC \NC \NR
\NC access       \NC \NC \NR
\NC change       \NC \NC \NR
\NC permissions  \NC \NC \NR
\NC nlink        \NC \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title=\type{pngdecode}]

This module is experimental and used in image inclusion. It is not some general
purpose module and is supposed to be used in a very controlled way. The
interfaces might evolve.

\starttabulate[|Tw(12em)|T|T|]
\DB name        \BC arguments                      \BC results \NC \NR
\TB
\NC applyfilter \NC (str,nx,ny,slice)              \NC string \NC \NR
\NC splitmask   \NC (str,nx,ny,bpp,bytes)          \NC string \NC \NR
\NC interlace   \NC (str,nx,ny,slice,pass)         \NC string \NC \NR
\NC expand      \NC (str,nx,ny,parts,xline,factor) \NC string \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title=\type{basexx}]

Some more experimental helpers:

\starttabulate[|Tw(12em)|T|T|]
\DB name      \BC arguments \BC results \NC \NR
\TB
\NC encode16  \NC (str[,newline])  \NC string \NC \NR
\NC decode16  \NC (str)            \NC string \NC \NR
\NC encode64  \NC (str[,newline])  \NC string \NC \NR
\NC decode64  \NC (str)            \NC string \NC \NR
\NC encode85  \NC (str[,newline])  \NC string \NC \NR
\NC decode85  \NC (str)            \NC string \NC \NR
\NC encodeRL  \NC (str)            \NC string \NC \NR
\NC decodeRL  \NC (str)            \NC string \NC \NR
\NC encodeLZW \NC (str[,defaults]) \NC string \NC \NR
\NC decodeLZW \NC (str[,defaults]) \NC string \NC \NR
\LL
\stoptabulate

\stopsection

\startsection[title={Multibyte \type {string} functions}]

The \type {string} library has a few extra functions, for example \libidx
{string} {explode}. This function takes upto two arguments: \type
{string.explode(s[,m])} and returns an array containing the string argument \type
{s} split into sub-strings based on the value of the string argument \type {m}.
The second argument is a string that is either empty (this splits the string into
characters), a single character (this splits on each occurrence of that
character, possibly introducing empty strings), or a single character followed by
the plus sign \type {+} (this special version does not create empty sub-strings).
The default value for \type {m} is \quote {\type { +}} (multiple spaces). Note:
\type {m} is not hidden by surrounding braces as it would be if this function was
written in \TEX\ macros.

The \type {string} library also has six extra iterators that return strings
piecemeal: \libidx {string} {utfvalues}, \libidx {string} {utfcharacters},
\libidx {string} {characters}, \libidx {string} {characterpairs}, \libidx
{string} {bytes} and \libidx {string} {bytepairs}.

\startitemize
\startitem
    \type {string.utfvalues(s)}: an integer value in the \UNICODE\ range
\stopitem
\startitem
    \type {string.utfcharacters(s)}: a string with a single \UTF-8 token in it
\stopitem
\startitem
    \type {string.characters(s)}: a string containing one byte
\stopitem
\startitem
    \type {string.characterpairs(s)}: two strings each containing one byte or an
    empty second string if the string length was odd
\stopitem
\startitem
    \type {string.bytes(s)}: a single byte value
\stopitem
\startitem
    \type {string.bytepairs(s)}: two byte values or nil instead of a number as
    its second return value if the string length was odd
\stopitem
\stopitemize

The \type {string.characterpairs()} and \type {string.bytepairs()} iterators
are useful especially in the conversion of \UTF16 encoded data into \UTF8.

There is also a two|-|argument form of \type {string.dump()}. The second argument
is a boolean which, if true, strips the symbols from the dumped data. This
matches an extension made in \type {luajit}. This is typically a function that
gets adapted as \LUA\ itself progresses.

The \type {string} library functions \type {len}, \type {lower}, \type {sub}
etc.\ are not \UNICODE|-|aware. For strings in the \UTF8 encoding, i.e., strings
containing characters above code point 127, the corresponding functions from the
\type {slnunicode} library can be used, e.g., \type {unicode.utf8.len}, \type
{unicode.utf8.lower} etc.\ The exceptions are \type {unicode.utf8.find}, that
always returns byte positions in a string, and \type {unicode.utf8.match} and
\type {unicode.utf8.gmatch}. While the latter two functions in general {\it
are} \UNICODE|-|aware, they fall|-|back to non|-|\UNICODE|-|aware behavior when
using the empty capture \type {()} but other captures work as expected. For the
interpretation of character classes in \type {unicode.utf8} functions refer to
the library sources at \hyphenatedurl {http://luaforge.net/projects/sln}.

Version 5.3 of \LUA\ provides some native \UTF8 support but we have added a few
similar helpers too: \libidx {string} {utfvalue}, \libidx {string} {utfcharacter}
and \libidx {string} {utflength}.

\startitemize
\startitem
    \type {string.utfvalue(s)}: returns the codepoints of the characters in the
    given string
\stopitem
\startitem
    \type {string.utfcharacter(c,...)}: returns a string with the characters of
    the given code points
\stopitem
\startitem
    \type {string.utflength(s)}: returns the length of the given string
\stopitem
\stopitemize

These three functions are relative fast and don't do much checking. They can be
used as building blocks for other helpers.

\stopsection

\startsection[title={Extra \type {os} library functions}]

The \type {os} library has a few extra functions and variables: \libidx {os}
{selfdir}, \libidx {os} {selfarg}, \libidx {os} {setenv}, \libidx {os} {env}, \libidx {os}
{gettimeofday}, \libidx {os} {type}, \libidx {os} {name} and \libidx {os}
{uname}, that we will discuss here. There are also some time related helpers in
the \type {lua} namespace.

\startitemize

% selfbin
% selfpath
% selfdir
% selfbase
% selfname
% selfcore

\startitem
    \type {os.selfdir} is a variable that holds the directory path of the
    actual executable. For example: \type {\directlua {tex.sprint(os.selfdir)}}.
\stopitem

\startitem
    \type {os.selfarg} is a table with the command line arguments.
\stopitem

\startitem
    \type {os.setenv(key,value)} sets a variable in the environment. Passing
    \type {nil} instead of a value string will remove the variable.
\stopitem

\startitem
    \type {os.env} is a hash table containing a dump of the variables and
    values in the process environment at the start of the run. It is writeable,
    but the actual environment is \notabene {not} updated automatically.
\stopitem

\startitem
    \type {os.gettimeofday} returns the current \quote {\UNIX\ time}, but as a
    float. Keep in mind that there might be platforms where this function is
    not available.
\stopitem

\startitem
    \type {os.type} is a string that gives a global indication of the class of
    operating system. The possible values are currently \type {windows}, \type
    {unix}, and \type {msdos} (you are unlikely to find this value \quote {in the
    wild}).
\stopitem

\startitem
    \type {os.name} is a string that gives a more precise indication of the
    operating system. These possible values are not yet fixed, and for \type
    {os.type} values \type {windows} and \type {msdos}, the \type {os.name}
    values are simply \type {windows} and \type {msdos}

    The list for the type \type {unix} is more precise: \type {linux}, \type
    {freebsd}, \type {kfreebsd}, \type {cygwin}, \type {openbsd}, \type
    {solaris}, \type {sunos} (pre-solaris), \type {hpux}, \type {irix}, \type
    {macosx}, \type {gnu} (hurd), \type {bsd} (unknown, but \BSD|-|like), \type
    {sysv}, \type {generic} (unknown). But \unknown\ we only provide \LUAMETATEX\
    binaries for the mainstream variants.

    Officially we only support mainstream systems: \MSWINDOWS, \LINUX, \FREEBSD\
    and \OSX. Of course one can build \LUAMETATEX\ for other systems, in which
    case on has to check the above.
\stopitem

\startitem
    \type {os.uname} returns a table with specific operating system
    information acquired at runtime. The keys in the returned table are all
    string values, and their names are: \type {sysname}, \type {machine}, \type
    {release}, \type {version}, and \type {nodename}.
\stopitem

\stopitemize

\stopsection

\startsection[title={The \type {lua} library functions}]

The \type {lua} library provides some general helpers.

\startitemize

\startitem
    The \type {newtable} and \type {newindex} functions can be used to create
    tables with space reserved beforehand for the given amount of entries.
\stopitem

\startitem
    The \type {getstacktop} function returns a number that can be used for
    diagnostic purposes.
\stopitem

\startitem
    The functions \type {getruntime}, \type {getcurrenttime}, \type
    {getpreciseticks} and \type {getpreciseseconds} return what their name
    suggests.
\stopitem

\startitem
    On \MSWINDOWS\ the \type {getcodepage} function returns two numbers, one
    for the command handler and one for the graphical user interface.
\stopitem

\startitem
    The name of the startup file is reported by \type {getstartupfile}.
\stopitem

\startitem
    The \LUA\ version is reported by \type {getversion}.
\stopitem

\startitem
    The \type {lua.openfile} function can be used instead of \type {io.open}. On
    \MSWINDOWS\ it will convert the filename to a so called wide one which means
    that filenames in \UTF8 encoding will work ok. On the other hand, names given
    in the codepage won't.
\stopitem

\stopitemize

\stopsection

\stopchapter

\stopcomponent
