% language=uk

\startcomponent mk-fallback

\environment mk-environment

\chapter {Virtual Reality}

When a font lacks glyphs we can add new ones by making the font
virtual. A virtual font has virtual glyphs: instead of a reference
to a slot in the current font, such a glyph refers to a slot in
another font, or it combines several glyphs into one, or it just
contains code that ends up in the result (for instance a sequence
of \PDF\ commands that describes the shape). For \TEX\ a character
and its dimensions are what matters and what ends up in the result
is mostly a matter for the backend. In \LUATEX\ the backend is
integrated but even then during the typesetting process only the
characteristics of a glyph are used and not the shape.

In \CONTEXT\ we have a feature called \quote {compose} which
extends the font with extra characters and constructs its
representation from those of other characters.

\starttyping
\definefontfeature
  [composes]
  [kern=yes,ligatures=yes,compose=yes]
\stoptyping

When this feature is applied, \CONTEXT\ will try to fill in the
gaps in the \UNICODE\ vector of the font based on for instance
(de)composition information. Of course this has some limitations.
For instance \OPENTYPE\ fonts can ships with features, like
smallcaps. Currently we ignore this property when we add composed
characters. Technically it is no big deal to add variants but we
simply didn't do it yet at the time of this writing. After all,
such fallbacks can best be avoided by using proper fonts.

Our \CONTEXT\ \MKIV\ wishlist mentions a mechanism for combining
fonts into one font. For this we can use virtual fonts and the
machinery for that is in available in \LUA\ code. However such a
mechanism will be used for more drastic completion of a font than
the compose feature. For instance, often Chinese fonts lack proper
Latin glyphs and vise versa. But when we combine such fonts we
really do want to keep \OPENTYPE\ features working and so we
cannot use virtual fonts (unless we start merging features which
can become really messy and runtime consuming).

There is a relative simple solution using real fonts that kind of
behave like virtual ones: virtual real fonts. The trick is in the
fact that \TEX\ permits access to characters not present in the
font. Say that we have

\starttyping
<char 123><char 124><char 125>
\stoptyping

and that slot~124 has no glyph. In that case \TEX\ just inserts a
glyph node with a reference to the current font and this
character. Of course, when we let \TEX\ carry on, at some point it
will need glyph properties like the width, height and/or depth.
And in the backend, when writing the result to file, \TEX\ wants
to insert the glyph data in the file. In both cases we end up with
a message in the log file and a result file with missing data.

In \CONTEXT\ \MKIV\ we intercept the node lists at several points
and one of those is directly after the construction. So let's
consider the previous example again.

\starttyping
<font 32 char 123><font 32 char 124><font 32 char 125>
\stoptyping

Because the font has no character 124 we need a way to substitute
it with another character. All we have to do is to change the font
identifier~32 into one that makes sense. Such a replacement loop
is kind of trivial.

\starttyping
for n in traverse_id(glyph,head) do
    local v = vectors[n.font]
    if v then
        local id = v[n.char]
        if id then
            n.font = id
        end
    end
end
\stoptyping

We have a table (\type{vectors}) that can have a subtable (\type
{v}) for font with id (\type {n.font}) in which there can be a
reference from the current character (\type {n.char}) to another
font (\type {id}) that we use to replace the font reference (\type
{n.font}).

Filling the table is relatively easy but an explanation is beyond
this chapter. We only show the high level interface, one that
certainly will evolve.

\starttyping
\definefontfallback
  [SerifFallback]
  [Mono]
  [0x000-0x3FF]
  [check=yes,force=no]
\stoptyping

This command registers an entry in the \type {SerifFallback}
namespace. There can be multiple replacement in row (by just using
more of these commands), but here we have only one. The range
0x000--0x3FF will be checked and if the main font lacks a glyph in
that range, it will be taken from the font with the symbolic name
\type {Mono}. That name will be resolved when the fallback is
associated with a font. The \type {check} option tells the
machinery that we need to check for existence and because we don't
\type {force}, we will only replace missing glyphs. There is also
an \type {rscale} option, that permits relative scaling of the
fallback font to the main font, something that may be needed when
fonts come from different sources.

\starttyping
\definefontsynonym
  [SerifPlus]
  [Serif]
  [fallbacks=SerifFallback]
\stoptyping

This command associates a fallback with a font. There is always a
parent font and that is the font that triggers the checking of the
node list.

\starttyping
\definefont [MySerif] [SerifPlus at 10pt]
\stoptyping

Here we defines a font called \type {\MySerif} that refers to a
symbolic name \type {SerifPlus} which in turn refers to the
current \type {Serif} font (these symbolic names are resolved in
typescripts, one of the building blocks of \CONTEXT s font
system). The mentioned fallbacks will be initialized when the font
is defined. This examples demonstrates that there is a clean
separation between font definitions and fallbacks. This makes it
possible to share fallback definitions.

So, let's summarize what happens:

\startitemize[packed]
\item a font is defined in the normal way but has falbacks
\item the associated fallback fonts are defined too
\item the main font gets a table with fallback id's
\item the main font is used in the document stream
\item the node list is intercepted and parsed for this font
\item references to fallback fonts take care of missing glyphs
\stopitemize

We end with an example.

\startbuffer
\definefontfallback [Demo] [Mono] [0x30-0x39] [force=yes]
\definefontsynonym  [DemoSerif] [Serif] [fallbacks=Demo]

\definefont [MyDemoSerif] [DemoSerif at 20pt]

\MyDemoSerif Here the digits, like 13579, are replaced.
\stopbuffer

\typebuffer

\start \blank[3*medium] \getbuffer \blank[3*medium] \stop

Beware: the fallback definitions are global, but this is hardly a
problem because normal such trickery is taking place at the document
level.

\stopcomponent
