% language=uk

\environment mk-environment

\startcomponent mk-nicetoknow

\chapter{Nice to know}

\section{Tricky ligatures}

Getting the 1.06 release of Latin Modern out in the wild took some discussion and
testing. Not only were the names (internal names as well as file names) changed in
such a way that multiple paplications could deal with it, but also some more
advanced ligature trickery was added.

\startbuffer
\definefontfeature
  [ijtest]
  [mode=node,
   script=latn,language=nld,strategy=3,
   liga=yes,kern=yes]

\definefont
  [ijfont]
  [name:lmroman10regular*ijtest at 36pt]

\start \ijfont \setstrut fijn ijsje fiets flink effe\stop
\stopbuffer

\typebuffer

This bit of Dutch shows up as:

\getbuffer

Do you see the trick? There are both an ij and an fi ligature, but we need to prevent
the ij ligature in fijn. Of course not all fonts have this feature, which indicated
that you can never depend on it.

\section {Herds}

A while ago, Duane, Taco and I published the Cow Font. It's non||trivial to
cook up a font made of cows, but of course Mojca Miklavec (who else) wants
to typeset something Slovenian in this font. Now, the problem is that in
\MKIV\ we don't have fallback characters, or more precisely, we don't make
\UTF\ characters active and accent composing commands are mapped onto \UTF.

This means that nothing will show up when we have no characters in the
defined fonts. For the moment we stick to simple virtual fonts but
because we can use node lists in virtual fonts, in the near future we will
cook up a way to create arbitrary fallback characters.

The following example demonstrates how to \quote {complete} a font that misses
glyphs.

\startbuffer
\definefontfeature[coward] [kern=yes,ligatures=yes]
\definefontfeature[cowgirl][kern=yes,ligatures=yes,compose=yes]

\definefontsynonym [cows] [koeieletters.afm*coward]
\definefontsynonym [herd] [koeieletters.afm*cowgirl]

\blank[3*medium]
\dontleavehmode\hbox{\definedfont[cows sa 5](č)(š)(ž)}
\blank[3*medium]
\dontleavehmode\hbox{\definedfont[herd sa 5](č)(š)(ž)}
\blank[3*medium]
\dontleavehmode\hbox{\definedfont[herd sa 5](\v{c})(\v{s})(\v{z})}
\stopbuffer

\typebuffer

As expected (at least by me) the first line has no compose characters.

\getbuffer

\stopcomponent
