\setuplayout
  [width=middle,
   height=middle,
   topspace=2cm,
   header=0pt,
   footer=1cm]

\setupbodyfont
  [bookman]

\usemodule
  [punk,abr-02]

\setuphead
  [section]
  [color=ColorThree,
   style=\bfb]

\setupitemgroup
  [itemize] [each]
  [packed] [color=ColorThree,symcolor=ColorThree]

\setupfooter
  [color=ColorThree,
   style=bold]

\setupfootertexts
  [pagenumber]

\setupwhitespace
  [big]

\definefont[PunkFont][demo@punk at 20pt]

% \def\aterm{\sym{?}}
% \def\rterm{\sym{--}}
% \def\dterm{\sym{+}}
% \def\pterm{\sym{!}}
%
% \startitemize[packed]
%     \aterm on the agenda (update, extension, rewrite)
%     \rterm no longer on the agenda, rejected
%     \dterm no longer on the agenda, done
%     \pterm work in progress (so keep an eye on the betas)
% \stopitemize

\definecolor[ColorOne]  [c=0.5,m=0.2,y=0.5,k=0.2]
\definecolor[ColorTwo]  [c=0.5,m=0.5,y=0.1,k=0.1]
\definecolor[ColorThree][c=0.1,m=1.0,y=1.0,k=0.2]

\starttext

\startMPpage
    StartPage ;
        numeric n ; n := 200 ;
        numeric o ; o :=  25 ;

        pair p[] ;
        for i=1 upto n :
            p[i] = (o + uniformdeviate (PaperWidth-2*o), o + uniformdeviate (PaperHeight-2*o)) ;
        endfor ;

        picture d ; d := image (      for i=1 upto n :             draw p[i] ; endfor ) ;
        picture l ; l := image ( draw for i=1 upto n : if i > 1 : -- fi p[i]   endfor ) ;
        picture t ; t := textext("\framed[frame=off,align={middle,lohi},foregroundcolor=ColorThree,foregroundstyle=\PunkFont]{\ConTeXt\endgraf MkIV\endgraf\kern-\strutdepth RoadMap}") ;

        fill Page enlarged 10 withcolor "ColorOne" ;

        draw d withcolor white      withpen pencircle scaled  o      ;
        draw d withcolor "ColorTwo" withpen pencircle scaled (o - 5) ;
        draw l withcolor white      withpen pencircle scaled (o / 5) ;
        draw l withcolor "ColorTwo" withpen pencircle scaled (o /10) ;
        draw thelabel.ulft((t xsized .5PaperWidth),lrcorner Page shifted - (PaperWidth/20,-PaperWidth/40)) ;
    StopPage ;
\stopMPpage

\startsubject[title={Introduction}]

There is not really a long term roadmap for development. One reason is that there
is already a lot available. When we started with \LUATEX, the \CONTEXT\ code was
mostly rewritten from \MKII\ to \MKIV, and that process is more of less finished.
Of course there is always work left. We have moved on to \LUAMETATEX\ and \LMTX\
as follow up on \MKIV. Deep down some things might (and will) change but the user
interface will stay the same (as usual).

This file is not a complete overview of our plans but users can at least get an
idea of what we're working on and what is coming. Feel free to submit
suggestions.

\startlines
Hans Hagen
Hasselt NL
\currentdate
\stoplines

\stopsubject

\startsubject[title={On the agenda for \LUAMETATEX}]

The agenda for \LUATEX\ is basically empty as this program is (supposed to be)
frozen, but for \LUAMETATEX\ we can still consider improvements. In fact, there
are quite some changes between the engines already (internally). Items that
concern \LUATEX\ have been removed from the agenda but some improvements in
\LUAMETATEX\ might trickle back.

\startitemize
    \startitem
        cleanup passive nodes in the par builder
    \stopitem
    \startitem
        optimize some callback resolution (more direct) but there is not that
        much to gain there
    \stopitem
    \startitem
        remove local par in head of line when done with linebreak and maybe also
        ensure leftskip and rightskip
    \stopitem
    \startitem
        only return nil in lua helpers when we expect multiple calls in in one
        line
    \stopitem
    \startitem
        experiment a bit more with the new protrusion code
    \stopitem
\stopitemize

\stopsubject

\startsubject[title={On the agenda for \CONTEXT\ \MKIV}]

\startitemize
    \startitem
        play with par callback and properties
    \stopitem
    \startitem
        optimize positions for columnareas and parpos (sequential)
    \stopitem
    \startitem
        add flag to font for math engine
    \stopitem
    \startitem
        get rid of components in glyph nodes
    \stopitem
    \startitem
        play with box attributes
    \stopitem
    \startitem
        check consistency between footnotes and running text (main color,
        styles, properties)
    \stopitem
    \startitem
        freeze actions and tasks (by name or function), maybe a register
        function that does that so no one can overload built-in features
    \startitem
        redo some of the spacing (adapt to improvements in engine)
    \stopitem
    \startitem
        more node and code injections
    \stopitem
    \startitem
        maybe reorganize position data (more subtables)
    \stopitem
    \startitem
        use \type {\matheqnogapstep}, \type {\Ustack}, \type {\mathscriptsmode},
        \type {\mathdisplayskipmode} and other new math primitives
    \stopitem
    \startitem
        cleanup some lua helpers (io.exists vs lfs.isfile)
    \stopitem
\stopitemize

\stopsubject

\stoptext
