\startcomponent ma-cb-en-columns

\enablemode[**en-us]

\project ma-cb

\startchapter[title=Columns]

\index{columns}

\Command{\tex{startcolums}}
\Command{\tex{setupcolumns}}
\Command{\tex{column}}

Simple sections of text can be typeset in columns. If you preceed a text fragment
by \type{\startcolumns} and close the text fragment by \type{\stopcolumns}
everything in between will be set in columns.

\shortsetup{startcolumns}

Let's give an example:

\startbuffer
\startcolumns[n=3,tolerance=verytolerant]
  Hasselt is an old Hanseatic City, situated 12~km north of Zwolle
  at the river Zwartewater.

  ...

  Furthermore some events of special interest should be
  mentioned. Every year at the end of August Hasselt celebrates
  the \quote{Eui Festival} (hay festival).
\stopcolumns
\stopbuffer

\typebuffer

\startbuffer
\hyphenation{Ste-pha-nus}
\startcolumns[n=3,tolerance=verytolerant]
Hasselt is an old Hanseatic City, situated 12~km north of Zwolle at
the river Zwartewater.

The city has a long history since obtaining the city charter around
1252. Part and parcel of this history can be traced back to a large
number of monuments to be admired in the city center.

There you will find the St. Stephanus church, a late gothic church
dating back to 1479 with a magnificent organ. The former Municipal
Building is situated on The Market Place. Constituted between 1500
and 1550 it houses a large collection of weapons, amongst which one
of the largest collection of black powder guns (haakhussen) in the
whole world should be mentioned.

Furthermore there is a corn windmill \quote{The Swallow}, dating back to
1748 as well as the \quote{Stenendijk}, a unique embankment and the last
shell limekiln in Europe still in full operation.

The city center with the townmoat adorned by lime-trees, the Van
Stolkspark and the hustle and bustle at the docks are ideally suited
for a stroll.

The area around Hasselt is also worth mentioning. In wintertime
polder Mastenbroek harbours large numbers of geese. In summertime the
hamlets Gen\-ne, Streukel and Celle\-mui\-den form, together with the very
rare lapwing flowers (Lat. Fritillaria meleagris) found on the banks
of the river Zwatewater, the ideal surroundings for walking or
cycling trips.

Hasselt also is a very important center for watersports. The lakes of
northwest Overijssel, the river IJssel, the Overijsselse Vecht and
the Randmeren are within easy reach from the yacht harbour `De
Molenwaard'. Sailing, fishing, swimming and ca\-noe\-ing can be fully
enjoyed in Hasselt.

Furthermore some events of special interest should be
mentioned. Every year at the end of August Hasselt celebrates the
\quote{Eui Festival} (hay festival).
\stopcolumns
\stopbuffer

The result will be a three column text.

\start
\switchtobodyfont[9pt]
\getbuffer
\stop

If possible a new column can be enforced with \type{\column}. You can set up
columns with:

\shortsetup{setupcolumns}

In most cases you will obtain a better result by type setting the text on
\quote{grid}. This is done by typing \type{grid=yes} in the command \type
{\setuplayout}.

If you want to use columns within a framed text \type{\start ... \stopframedtext}
there is the simple column mechanism.

\startbuffer
\startframedtext[background=color,backgroundcolor=gray]
\startsimplecolumns
  In Hasselt's local newspaper there was a column on the
  local customs during New Years Eve.
  ...
  \midaligned{\inlinechemical{CaC_2,+,2H_2O,GIVES,C_2H_2(g),+,Ca(OH)_2}}
  ...
  Nowadays the heavy metal lid of the milk can is replaced by
  a football. This does not reduce the sound but it is much
  saver.
\stopsimplecolumns
\stopbuffer

\typebuffer

\startbuffer
\startframedtext[background=color,backgroundcolor=gray,width=\textwidth]
\startsimplecolumns
  In Hasselt's local newspaper there was a column on the
  local customs during new years Eve. Next to the more
  general custom of eating Dutch doughnuts (oliebollen)
  and lighting fireworks there is the carbide shooting.
  What you need is an oldfashioned metal milk can, carbide,
  a little water and a lighter.

  The carbide and water is mixed in the closed milk can and
  will produce \inlinechemical{C_2H_2} gas (acetylene), via:

  \midaligned{\inlinechemical{CaC_2,+,2H_2O,GIVES,C_2H_2 (g),+,Ca(OH)_2}}

  The volatile acetylene gas in the milk can is ignited via a small
  opening in the can. The result is a very loud detonation and the lid
  flies off.

  It will not surprise you that Hasselts youth has a designated
  shooting ground for carbide shooting. Nowadays the heavy
  metal lid of the milk can is replaced by a football. This
  does not reduce the sound but it is much saver!
\stopsimplecolumns
\stopframedtext
\stopbuffer

This will result in:

\getbuffer

There is an advanced column mechanism available that is described in the
\goto {\em Columns} [ url (manual:columns)
] manual.

\stopchapter

\stopcomponent
