\startcomponent ma-cb-en-setupcommands

\enablemode[**en-us]

\project ma-cb

\startchapter[title=Setup commands]

\index{set up}
\index{layout}

Global commands are placed in the setup area of your input file, before
\type{\starttext}. In \in{appendix}[commandsetups] there is a complete overview of the
available commands and their parameters.

The set up commands all have the same structure. They look something like:

\shortsetup{setupparagraphs}

A set up command consists of a more or less logical name and a number of bracket
pairs. Bracket pairs may be optional and in that case the \type{[]} are typeset
slanted {\tt \sl []}. In the definition the bracket pairs may contain:

\starttyping
\setupacommand[.1.][.2.][..,..=..,..]
\stoptyping

The commas indicate that a list of parameters can be enclosed. In the options
list following the definition, the \type{.1.} and \type{.2.} show the possible
options that can be set in the first and second bracket pair respectively. The
parameters and their possible values are placed in the third bracket pair.

The default options and parameter values are underlined.

Furthermore you will notice that some values are typeset in a slanted way: {\sl
section}, {\sl name}, {\sl dimension}, {\sl number}, {\sl command} and {\sl
text}. This indicates that you can set the value yourself.

\starttabulate[|S||]
\NC section   \NC a section name like chapter, section, subsection etc. \NC\NR
\NC name      \NC an identifier (logical name)                          \NC\NR
\NC dimension \NC a dimension with a unit in \type{cm}, \type{pt},
                  \type{em}, \type{ex}, \type{sp} or \type{in}          \NC\NR
\NC number    \NC an integer                                            \NC\NR
\NC command   \NC a command                                             \NC\NR
\NC text      \NC text                                                  \NC\NR
\stoptabulate

In the \goto {Quick Reference} [ url (manual:qr-en) ]
manual you can find a complete overview of the commands and their parameters.

\stopchapter

\stopcomponent
