\environment publications-style

\startcomponent publications-tracing

\startchapter
  [title=Tracing]

There are several tracing options. If you want to see where a citations refers to
and where a list entry point back to, you can say:

\startTEX
\enabletrackers[publications.crosslinks]
\stopTEX

This injects markers in both places. One list entry can point to multiple
citations. The other tracers a more for debugging and can generate lots of
messages.

\startTEX
publications
publications.authorhash
publications.cite
publications.cite.missing
publications.cite.references
publications.detail
publications.duplicates
publications.match
publications.setups
publications.sorters
publications.suffixes
\stopTEX

You can also trace the databases. Take this one:

\startbuffer
\startbuffer[phony]
@comment { warning :
    Hello Allan!
    How are you doing today?
}

@CoMMeNT { message :
    Hello Allan!
    How are you doing today?
}

@comment {
    all kind of crap
}

@Article{Myarticle,
  Title                    = {My title},
  Author                   = {Myself, Me},
  Journal                  = {My favorite journal},
  Year                     = {2014},
  Pages                    = {1},
}

@Foo{foo,
  Goo = goo,
}
\stopbuffer
\stopbuffer

\typeBTXbuffer \getbuffer

\startbuffer
\usebtxdataset [phony] [phony.buffer]
\stopbuffer

When we load this database (buffer) with:

\typeTEXbuffer \getbuffer

We get this on the console and in the log

\starttyping
publications > adding bib data to set 'phony' from source 'phony'
publications > phony > warning : Hello Allan!
publications > phony > warning : How are you doing today?
publications > phony > message : Hello Allan!
publications > phony > message : How are you doing today?
\stoptyping

You can use this feature to add warnings to your database for entries that you
need to check.

You can also use comment to hide entries:

\startBTX
@comment {

    @article{Hobby1999,
      author  = {Hobby, John D.},
      year    = {1999},
      title   = {Introduction to MetaPost},
      journal = {Eutupon},
      volume  = {2},
      month   = {April},
      pages   = {39-53},
    }

}
\stopBTX

\stopchapter

\stopcomponent
