% language=us

\startcomponent followingup-compilation

\environment followingup-style

\logo[WLS]       {WLS}
\logo[INTEL]     {Intel}
\logo[APPLE]     {Apple}
\logo[UBUNTU]    {Ubuntu}
\logo[RASPBERRY] {RaspberryPi}

\startchapter[title={Compilation}]

Compiling \LUATEX\ is possible because after all it's what I do on my machine.
The \LUATEX\ source tree is part of a larger infrastructure: \TEX Live. Managing
that one is work for specialists and the current build system is the work of
experts over a quite long period of time. When you only compile \LUATEX\ it goes
unnoticed that there are many dependencies, some of which are actually unrelated
to \LUATEX\ itself but are a side effect of the complexity of the build
structure.

When going from \LUATEX\ to \LUAMETATEX\ many dependencies were removed and I
eventually ended up with a simpler setup. The source tree went down to less than
30 MB and zipped to around 4 MB. That makes it possible to consider adding the
code to the regular \CONTEXT\ distribution.

One reason for doing that is that one keeps the current version of the engine
packaged with the current version of \CONTEXT. But a more important one is that
it fulfils a demand. Some time ago we were asked by some teachers participating
in a (basically free) math method for technical education what guarantees there
are that the tools used are available forever. Now, even with \LUAMETATEX\ one
has to set up a compiler but it is much easier than installing the whole \TEX
Live infrastructure for that. A third reason is that it gives me a comfortable
feeling that I myself can compile it anywhere as can \CONTEXT\ users who want to
do that.

The source tree traditionally has libs in a separate directory (lua, luajit, zlib
and zziplib). However, it is more practical to have them alongside our normal
source. These are relative small collections of files that never change so there
is no reason not to do it. \footnote {If I ever decide to add more libraries,
only the minimal interfaces needed will be provided, but at this moment there are
no such plans.}

Another assumption we're going to make is that we use 64 bit binaries. There is
no need to support obsolete platforms either. As a start we make sure it compiles
on the platforms used by \CONTEXT\ users. Basically we make a kind of utility.
For now I can compile the \WINDOWS\ 32 bit binaries that my colleague needs in
half a minute anyway, but in the long run we will settle for 64 bits.

I spent about a week figuring out why the compilation is so complex (by
selectively removing components). At some point compilation on \OSX\ stopped
working. When the minimum was reached I decided to abandon the automake tool
chain and see if \type {cmake} could be used (after all, Mojca challenged that).
In retrospect I should have done that sooner because in a day I could get all
relevant platforms working. Flattening the source tree was a next step and so
there is no way back now. What baffled me (and Alan, who at some point joined in
testing \OSX) is the speed of compilation. My pretty old laptop needed about half
a minute to get the job done and even on a \RASPBERRY\ with only a flash card
just a few minutes were needed. At that point, as we could remove more make
related files, the compressed 11 MB archive (\type {tar.xz}) shrunk to just over
2~MB. Interesting is that compiling \MPLIB\ takes most time, and when one compiles
in parallel (on more cores) that one finishes last.

For the record: I do all this on a laptop running \MSWINDOWS\ 10 using the Linux
subsystem. When that came around, Luigi made me a working setup for cross
compilation but in the meantime with GCC 8.2 all works out of the box. I edit the
files at the \MSWINDOWS\ end (using \SCITE), compile at the \LINUX\ end, and test
everything on \MSWINDOWS. It is a pretty convenient setup.

When compilation got faster it became also more convenient to do some more code
reshuffling. This time I decided to pack the global variables into structures,
more or less organized the way the header files were organized. It gives a bit
more verbosity but also has the side effects that (at least in principle) the
\CPU\ cache can perform better because neighboring variables are often cached as
part of the deal. Now it might be imagination, but in the process I did notice
that mid March processing the manual went down to below 11.7 seconds while before
it stayed around 12.1 seconds. Of course this is not that relevant currently, but
I might make a difference on less capable processors (as in a low power setup).
It anyway didn't hurt.

In the meantime some of the constants used in the program got prefixes or
suffixes to make them more unique and for instance the use of \type {normal} as
equivalent for zero was made a bit more distinctive as we now have more subtypes.
That is: all the subtypes were collected in enumerations instead of \CCODE\
defines. Back to the basics.

End of 2020 I noticed that the binary had grown a bit relative to the mid 2020
versions. This surprised me because some improvements actually made them smaller,
something you notice when you compile a couple of times when doing these things.
I also noticed that the platforms on the compile farm had quite a bit of
variation. In most cases we're still below my 3MB threshold, but when for
instance cross compiled binaries become a few hundred MB larger one can get
puzzled. In the \LUAMETAFUN\ manual I have this comment at the top:

\starttyping[style=\ttx]
------------------------   ------------------------   ------------------------
2019-12-17  32bit  64bit   2020-01-10  32bit  64bit   2020-11-30  32bit  64bit
------------------------   ------------------------   ------------------------
freebsd     2270k  2662k   freebsd     2186k  2558k   freebsd     2108k  2436k
openbsd6.6  2569k  2824k   openbsd6.6  2472k  2722k   openbsd6.8  2411k  2782k
linux-armhf 2134k          linux-armhf 2063k          linux-armhf 2138k  2860k
linux       2927k  2728k   linux       2804k  2613k   linux   (?) 3314k  2762k
                                                      linux-musl  2532k  2686k
osx                2821k   osx                2732k   osx                2711k
ms mingw    2562k  2555k   ms mingw    2481k  2471k   ms mingw    2754k  2760k
                                                      ms intel           2448k
                                                      ms arm             3894k
                                                      ms clang           2159k
------------------------   ------------------------   ------------------------
\stoptyping

So why the differences? One possible answer is that the cross compiler now uses
\GCC9 instead of \GCC8. It is quite likely that inlining code is done more
aggressively (at least one can find remarks of that kind on the Internet). An
interesting exception in this overview is the \LINUX\ 32 bit version. The native
\WINDOWS\ binary is smaller than the \MINGW\ binary but the \CLANG\ variant is
still smaller. For the native compilation we always enabled link time
optimization, which makes compiling a bit slower but similar to regular
compilation in \WLS\ but when for the other compilers we turn on link time
optimization the linker takes quite some time. I just turn it off when testing
code because it's no fun to wait these additional minutes with \GCC. Given that
the native windows binary by now runs nearly as fast as the cross compiled ones,
it is an indication that the native \WINDOWS\ compiler is quite okay. The numbers
also show (for \WINDOWS) that using \CLANG\ is not yet an option: the binaries
are smaller but also much slower and compilation (without link time optimization)
also takes much longer. But we'll see how that evolves: the compile farm
generates them all.

So, what effects does link time optimization has? The (current) cross compiled
binary is is some 60KB smaller and performs a little better. Some tests show
some 3~percent gain but I'm pretty sure users won't notice that on a normal run.
So, when we forget to enable it when we release new binaries, it's no big deal.

Another end 2020 adventure was generating \ARM\ binaries for \OSX\ and \WINDOWS.
This seems to work out well. The \OSX\ binaries were tested, but we don't have
the proper hardware in the compile farm, so for now users have to use \INTEL\
binaries on that hardware. Compiling the \LUAMETATEX\ manual on a 2020 M1 is a
little more that twice as fast than on my 2013 i7 laptop running \WINDOWS. A
native \ARM\ binary is about three times faster, which is what one expects from a
more modern (also a bit performance hyped) chipset. On a \RASPBERRY\ with 4MB
ram, an external \SSD\ on \USB3, running \UBUNTU\ 20, the manual compiles three
times slower than on my laptop. So, when we limit conclusions to \LUAMETATEX\ it
looks like \ARM\ is catching up: these modern chipsets (from \APPLE\ and
\MICROSOFT, although the later was not yet tested) with plenty of cache, lots of
fast memory, fast graphics and speedy disks are six times faster than a cheap
media oriented \ARM\ chipset. Being a single core consumer, \LUAMETATEX\ benefits
more from faster cores than from more cores. But, unless I have these machines
on my desk these rough estimates have to do.

\stopchapter

\stopcomponent
