% language=us

\startcomponent followingup-feedback

\environment followingup-style

\logo [AMD]   {AMD}
\logo [INTEL] {Intel}

\startchapter[title={Feedback}]

\startsection[title={Introduction}]

As \LUATEX\ 1.10 is basically frozen in terms of functionality not much can or
will be added. But it made sense to some of the (small) improvements that were
made in \LUAMETATEX\ got feedback to \LUATEX\ (or will be at some point). Because
we are also experimenting, there can be a delay. \footnote {Later chapters
mention a few more possible extensions.}

Of course the question is \quotation {Should we feedback (retrofit) at all?}. I'm
still not sure about it. There should be a good reason to do it because it can
harm stability of the original. At some point \CONTEXT\ can default to the follow
up in which case testing the original becomes more difficult for specific
features. I never ran into (useful) demands for extensions so retrofit can have a
low priority.

Another factor is that when distributions start adding stuff to stock \LUATEX\ on
top of what is our default (after all isn't that what happens with open source
projects), it makes not much sense to look back and retrofit new functionality,
because there is not much change that we will use such a variant ourselves and we
could introduce errors in the process. Providing bloatware is not our objective.

Related to this is the question if we should always go into \LMTX\ mode and I'm
no longer sure if we shouldn't do that. We can use plain \TEX\ with the regular
\LUATEX\ backend and just forget about some generic framework. The danger of it
backfiring is just too large. It is a waste of time and will keep us back.

One reason for a dual mode is that it made possible some timings in order to
determine bottlenecks. I did some rough tests and that is enough to get the
picture. Take this document:

\starttyping
\starttext
    \dorecurse
        {1000}
        {\samplefile{sapolsky} {\bf\samplefile{sapolsky}}\par}
\stoptext
\stoptyping

Using regular \LUATEX\ this takes on an \INTEL\ i7-3840 mobile processor about
9.3 seconds while \LUAMETATEX\ needs 11.2 seconds, so we loose time. This is
because we have only text so the native backend wins on piping out the page
stream. On my domotica fitlet with an low power \AMD\ A10 processor running
\LINUX\ the runtime goes from 25.4 seconds to 27.8 seconds, so again a slow down.

But this is not a typical document! Add a bit more spice and the numbers reverse.
For processing the \LUATEX\ manual stock \LUATEX\ takes 12.6 seconds on the
\INTEL\ and \LUAMETATEX\ needs 12.4 seconds. On the \AMD\ runtime goes from 35.1
seconds down to 32.8 seconds. So here we win some.

These are rough timings and a few weeks later we go these timings on the \INTEL:
\footnote {On the more modern gaming laptop of a nephew we measured half these
numbers.}

\starttabulate[|l|c|c|c|]
\BC engine            \BC backend \BC runtime \BC \LUAJIT\ vm \BC \NR
\HL
\NC \LUATEX\     1.10 \NC normal  \NC 12.4    \NC 9.9 \NC \NR
\NC \LUATEX\     1.10 \NC lmtx    \NC 12.7    \NC 9.8 \NC \NR
\NC \LUAMETATEX\ 2.00 \NC lmtx    \NC 12.2    \NC 9.3 \NC \NR
\stoptabulate

Because we have more \LUA\ code active, we pay a price with \LMTX\ but not on
\LUAMETATEX\ (as of now, later we will see a performance bump). The gain when
using the \LUAJIT\ virtual machine is more noticeable. And, there is probably
some more to gain. In case you wonder why this matters: think of the low power
\AMD\ processor. When we have to replace computers we can consider using low
power ones, with weaker processors, less memory, and less cache. For the record:
I use cross compiled \MINGW\ binaries on windows (they are quite a bit faster
than native windows binaries). And the binaries are less than 3MB (small files
and less resources are nice when running on remote file systems).

This all indicates that we have no real reason to stick to a mixed approach: if we
want we can just switch to always \LMTX\ and never look back.

\stopsection

\startsection[title={Expressions}]

When writing an article that involved using a \type {\numexpr} it struck me that
we should have a proper integer division. Just compare these:

\startbuffer[1]
\the\numexpr 13/2\relax
\stopbuffer

\startbuffer[2]
\scratchcounter13 \divide\scratchcounter 2 \the\scratchcounter
\stopbuffer

\typebuffer[1]

and

\typebuffer[2]

This gives {\bf \inlinebuffer[1]} and {\bf \inlinebuffer[2]}. We now also have:

\startbuffer[3]
\the\numexpr 13:2\relax
\stopbuffer

\typebuffer[3]

which gives {\bf \inlinebuffer[3]}. I considered using a double slash (as in
\LUA) but using a colon is easier. Of course those who make that an active
character are probably toast. This is an easy patch but it's hard to predict
possible side effects outside \CONTEXT.

\stopsection

\startsection[title={Looking ahead}]

Sometimes you want to look ahead and act upon the presence of a specific
character. Implementing that in pure \TEX\ primitives is no big deal, but
especially when you want to ignore leading spaces it leads to rather verbose code
when tracing is enabled. Out of curiosity I played with a primitive that can help
us out. Although there is also a performance gain, in practice one will not
notice that unless such a feature is used millions of times, but in that case the
gain is noise compared to the rest of the run.

\startbuffer
\def\foo{\futureexpand/\fooyes\foonop}
\def\fooyes/#1/#2{[#1:#2]}
\def\foonop    #1{(#1)}

\foo/yes/{one}\quad
\foo     {two}
\stopbuffer

\typebuffer

We either expand \type {\fooyes} or \type {\foonop}, depending on the presence
of a \type {/} after \type {\foo}. So, the result is:

{\tttf\getbuffer}

The next examples demonstrates two variants. The second one doesn't inject spaces
back into the stream.

\startbuffer
\def\f<{\futureexpand/\y\n}
\def\y/#1/{#1}
\def\n    {}

(\f</yes/>)\quad
(\f<     >)
\stopbuffer

\typebuffer

Watch the space in the \type {\n} case.

{\tttf\getbuffer}

\startbuffer
\def\f<{\futureexpandis/\y\n}
\def\y/#1/{#1}
\def\n    {}

(\f</yes/>)\quad
(\f<     >)
\stopbuffer

\typebuffer

This time the space is not injected (\type{is} is short for ignore spaces).

{\tttf\getbuffer}

I will probably use this one in \CONTEXT, but as said, not for performance
reasons but because it reduces code and therefore tracing. \footnote {In the
\CONTEXT\ code base there are several places where less code takes precedence
over efficiency. But in all cases extensive tests were done to see if it made a
dent in practical performance.}

\stopsection

\startsection[title={Checking numbers an dimensions}]

The \CONTEXT\ user interface often uses parameters that take keywords as well as
a number or dimension. In practice it has never been an issue to check for that but
there are a few cases where we'd like to be a bit more flexible.

\startbuffer
\doifelsenumber{123999999999999999}YN
\doifelsenumber               {123}YN
\doifelsenumber                 {A}YN
\doifelsenumber          {\char123}YN
\doifelsenumber          {\toks123}YN
\doifelsenumber{123\scratchcounter}YN

\doifelsedimension{123999999999999999pt}YN
\doifelsedimension               {123pt}YN
\doifelsedimension                   {A}YN
\doifelsedimension            {\char123}YN
\doifelsedimension            {\toks123}YN
\doifelsedimension    {123\scratchdimen}YN
\stopbuffer

\typebuffer

This typesets:

\startpacked
\getbuffer
\stoppacked

especially the \type {123\scratch...} checking is tricky. For that reason we now
have two new built|-|in checkers. Again, performance is not one of the arguments,
because these checks are not much faster than what we have already, they are just
a bit more robust for weird cases. A possible use of the primitives is:

\starttyping
\ifdimen123\or
  yes
\else % or \or
  no
\fi
\stoptyping

and:

\starttyping
\ifnumber123\or
  yes
\else % or \or
  no
\fi
\stoptyping

When a valid number or dimension is gobbled, the value pushed in the branches
is~1, and when an error is encountered the value~2 is pushed. Deep down we have
just an \type {\ifcase} and by not using the value zero we nicely skip the
invalid code. It might look a bit weird but we need a sentinel for the number
(and the \type {\or} serves as such, without introducing strange new constructs.
We'll see if we keep it (as testing must prove its usefulness).

\stopsection

\stopsection

\startsection[title={Comparing tokens}]

The following code compares (for instance) two strings:

\starttyping
\def\thisorthat#1#2%
  {\edef\one{#1}
   \edef\two{#2}
   \ifx\one\two
     this%
   \else
     that%
   \fi}

\thisorthat{foo}{bar}
\stoptyping

but this looks a bit cleaner (in a trace):

\starttyping
\def\thisorthat#1#2%
  {\iftok{#1}{#2}%
     this%
   \else
     that%
   \fi}

\thisorthat{foo}{bar}
\stoptyping

It's not that much faster (unless one uses it a real lot) as similar things have
to happen to get the test to work. But the nice things of this checker is that it
works with token registers and macros too. But in order use it in relevant places
in \CONTEXT\ I would have to adapt quite some code. This could actually be a
reason for a \MKIV\ freeze and \LMTX\ code base (as with \MKII). The question is:
does it pay off?

\stopsection

\stopchapter

\stopcomponent

% \newbox\mybox \setbox\mybox\ruledhbox{\blackrule[width=3cm,height=1cm]}

% \dontleavehmode\copy\mybox\quad \ruledhbox{\copy\mybox} \blank[3*line]
% \boxxmove\mybox            10pt
% \dontleavehmode\copy\mybox\quad \ruledhbox{\copy\mybox} \blank[3*line]
% \wd\mybox\dimexpr\wd\mybox+10pt\relax
% \dontleavehmode\copy\mybox\quad \ruledhbox{\copy\mybox} \blank[3*line]
% \boxymove\mybox            10pt
% \dontleavehmode\copy\mybox\quad \ruledhbox{\copy\mybox} \blank[3*line]
% \ht\mybox\dimexpr\ht\mybox+10pt\relax
% \dontleavehmode\copy\mybox\quad \ruledhbox{\copy\mybox} \blank[3*line]
