% language=uk

\startcomponent fonts-appendix

\environment fonts-environment

\startchapter[title=Appendix][color=darkgray]

\startsection[title=The \type {tfm} file,reference=fontdata:tfm]

The (binary) \type {tfm} file is not human readable but can be turned into a
verbose property list which is not that hard to understand.

\starttyping
tftopl texnansi-lmr10.tfm
\stoptyping

Here is an excerpt from the data file. It starts with some general properties of
the font. The \type {O} means that the value is in octal while the \type {R} is a
real. Keep in mind that \TEX\ has no datatype \quote {real} so internally it is
just integers representing scaled points.

\startnarrowtyping
(FAMILY LMROMAN10)
(FACE O 352)
(CODINGSCHEME LY1 ENCODING /TEX'N'ANSI, Y&Y/)
(DESIGNSIZE R 10.0)
(COMMENT DESIGNSIZE IS IN POINTS)
(COMMENT OTHER SIZES ARE MULTIPLES OF DESIGNSIZE)
(CHECKSUM O 4720464277)
\stopnarrowtyping

A text font has the following font dimensions:

\startnarrowtyping
(FONTDIMEN
    (SLANT      R 0.0)
    (SPACE      R 0.333333)
    (STRETCH    R 0.166667)
    (SHRINK     R 0.111112)
    (XHEIGHT    R 0.43055)
    (QUAD       R 1.0)
    (EXTRASPACE R 0.111112)
    ...
)
\stopnarrowtyping

Kerns and ligatures are packed into a table that is basically a sequence of
labelled entries. Here we see the entry for the character \type {f} which has
three ligatures: \type {ff}, \type {fi} and \type {fl}. Because ligatures can be
chained, octal slot 13 will have ligature entries for \type {ffl} and \type
{ffi}.

\startnarrowtyping
(LIGTABLE
    ...
    (LABEL C f)
    (LIG C f   O 13)
    (LIG C i   O 14)
    (LIG C l   O 10)
    (KRN O 135 R 0.027779)
    (KRN O  41 R 0.027779)
    (KRN O  51 R 0.027779)
    (KRN O  77 R 0.027779)
    (KRN O 223 R 0.027779)
    (KRN O 224 R 0.027779)
    (KRN O 140 R 0.027779)
    (KRN O  47 R 0.027779)
    (STOP)
    ...
)
\stopnarrowtyping

Each character gets its own entry. In this case there is no depth involved so it
is not shown. The comment is just a repetition of the entry in the ligtable.

\startnarrowtyping
(CHARACTER C f
    (CHARWD R 0.30555)
    (CHARHT R 0.688875)
    (CHARIC R 0.079222)
    (COMMENT
          (LIG C f   O 13)
          (LIG C i   O 14)
          (LIG C l   O 10)
          (KRN O 135 R 0.027779)
          (KRN O  41 R 0.027779)
          (KRN O  51 R 0.027779)
          (KRN O  77 R 0.027779)
          (KRN O 223 R 0.027779)
          (KRN O 224 R 0.027779)
          (KRN O 140 R 0.027779)
          (KRN O  47 R 0.027779)
    )
)
\stopnarrowtyping

\stopsection

\startsection[title=The \type {vf} file,reference=fontdata:vf]

A virtual font specification file can be converted to a more readable format with
\type {vftovp}, for instance:

\starttyping
vftovp eurm10.vf
\stoptyping

The information in a \type {vf} file will be combined with the data in the
accompanying \type {tfm} file so the output looks similar:

\startnarrowtyping
(VTITLE )
(FAMILY UNSPECIFIED)
(FACE F MRR)
(CODINGSCHEME TEX MATH ITALIC)
(DESIGNSIZE R 10.0)
(COMMENT DESIGNSIZE IS IN POINTS)
(COMMENT OTHER SIZES ARE MULTIPLES OF DESIGNSIZE)
(CHECKSUM O 24401046203)
(SEVENBITSAFEFLAG TRUE)
\stopnarrowtyping

Because this font is a math font there is no space defined.

\startnarrowtyping
(FONTDIMEN
    (SLANT      R 0.0)
    (SPACE      R 0.0)
    (STRETCH    R 0.0)
    (SHRINK     R 0.0)
    (XHEIGHT    R 0.459)
    (QUAD       R 1.0)
    (EXTRASPACE R 0.0)
)
\stopnarrowtyping

A virtual font will take glyphs from another font and therefore there are entries
that refer to these fonts. In the following definition index \type {0} is created
(the \type {D} specifies a decimal entry).

\startnarrowtyping
(MAPFONT D 0
    (FONTNAME     eurm10)
    (FONTCHECKSUM O 4276740471)
    (FONTAT       R 1.0)
    (FONTDSIZE    R 10.0)
)
(MAPFONT D 1
    (FONTNAME     cmmi10)
    (FONTCHECKSUM O 1350061076)
    (FONTAT       R 1.0)
    (FONTDSIZE    R 10.0)
)
\stopnarrowtyping

The zero indexed font is the default, so in the following entry this font is
taken:

\startnarrowtyping
(CHARACTER C W
    (CHARWD R 0.986)
    (CHARHT R 0.691)
    (CHARIC R 0.056)
    (COMMENT
        (KRN O 177 R  0.056)
        (KRN O  75 R -0.056)
        (KRN O  73 R -0.083)
        (KRN O  72 R -0.083)
    )
    (MAP
        (SETCHAR C W)
    )
)
\stopnarrowtyping

The next specification is a combination of two other glyphs becoming a new
glyph. We see here that the \type {MAP} table is actually a sort of program:

\startnarrowtyping
(CHARACTER O 200
    (CHARWD R 0.622)
    (CHARHT R 0.691)
    (MAP
        (PUSH)
        (MOVEDOWN   R -0.18)
        (MOVERIGHT  R  0.015)
        (SELECTFONT D  2)
        (SETCHAR    O 40)
        (POP)
        (SELECTFONT D  0)
        (SETCHAR    C  h)
    )
)
\stopnarrowtyping

The character information is also in the \type {tfm} companion and that is what
\TEX\ uses. The virtual information kicks in when the backend is creating the
page stream and embedding the fonts.

\stopsection

\startsection[title=The \type {map} file,reference=fontdata:map]

In a map file each line maps a font name onto a file that contains the font
shapes in bitmap or outline format. For instance in the file \type
{lm-texnansi.map} we find the line:

\startnarrowtyping
texnansi-lmr10 LMRoman10-Regular "enclmtexnansi ReEncodeFont" <lm-texnansi.enc <lmr10.pfb
\stopnarrowtyping

The backend will fetch the glyph data from \type {lmf10.pfb} and use the given
encoding file to resolve indices to glyph names. A \type {pfb} file can contain
more than 256 entries so names are used to access the data. The string between
quotes is used for the encoding vector in the resulting file.

The second entry in the line is the font name that will be used. This name is
also used to control subset behaviour. Multiple references to this name will be
collapsed into one inclusion when possible, thereby making the file as small as
possible. You better make sure that the names are unique for a specific font.

In addition to this, there can be directives for extending the font (horizontal
stretch) and transforming it into a slanted variant. Both are to be used with
care.

In \MKIV\ map files are only used for virtual math fonts and just as in \MKII\ we
load such files selectively. Users don't have to worry about this.

\stopsection

\startsection[title=The \type {enc} file,reference=fontdata:enc]

For historic reasons, an encoding file is a blob of \POSTSCRIPT, probably because
it can be copied into the final output directly. Given that \TEX\ got extended
anyway, you can wonder why this information never ended up in an extended \type
{tfm} or \type {vf} file. It had definitely made the traditional process much
more robust.

\startnarrowtyping
/enclmtexnansi[
/.notdef
/Euro
...
/dotaccent
/hungarumlaut
/ogonek
...
/ffi
/ffl
/dotlessi
/dotlessj
/grave
...
/thorn
/ydieresis
] def
\stopnarrowtyping

There are excactly 256 entries in such a vector and the names should match those
in a \type {pfb} file.

\stopsection

\startsection[title=The \type {afm} file,reference=fontdata:afm]

Here we show an excerpt from an \type {afm} file that comes with Latin Modern
Roman. Just as with a \type {tfm} file we start with some general information.
However we don't need to convert the file as is it already in human readable
format.

\startnarrowtyping
StartFontMetrics 2.0
Comment Generated by MetaType1 (a MetaPost-based engine)
Comment Creation Date: 7th October 2009
Notice Copyright 2003--2009 by B. Jackowski and J.M. Nowacki (on behalf of TeX USERS GROUPS).
Comment Supported by CSTUG, DANTE eV, GUST, GUTenberg, NTG, and TUG.
Comment METATYPE1/Type 1 version by B. Jackowski & J. M. Nowacki
Comment from GUST (http://www.gust.org.pl).
Comment This work is released under the GUST Font License.
Comment For the most recent version of this license see
Comment This work has the LPPL maintenance status `maintained'.
Comment The Current Maintainer of this work is Bogus\l{}aw Jackowski and Janusz M. Nowacki.
Comment This work consists of the files listed in the MANIFEST-Latin-Modern.txt file.
FontName LMRoman10-Regular
FullName LMRoman10-Regular
FamilyName LMRoman10
Weight Normal
ItalicAngle 0
IsFixedPitch false
UnderlinePosition -146
UnderlineThickness 40
Version 2.004
EncodingScheme FontSpecific
FontBBox -430 -290 1417 1127
CapHeight 683.33333
XHeight 430.55556
Descender -194.44444
Ascender 694.44444
Comment PFM parameters: LMRoman10 0 0 0xEE
Comment TFM designsize: 10 (in points)
Comment TFM fontdimen  1: 0          (slant)
Comment TFM fontdimen  2: 3.33333    (space)
Comment TFM fontdimen  3: 1.66667    (space stretch)
Comment TFM fontdimen  4: 1.11111    (space shrink)
Comment TFM fontdimen  5: 4.3055     (xheight)
Comment TFM fontdimen  6: 10         (quad)
Comment TFM fontdimen  7: 1.11111    (extra space)
Comment TFM fontdimen  8: 6.833      (non-standard: uc height)
Comment TFM fontdimen  9: 6.9445     (non-standard: ascender)
Comment TFM fontdimen 10: 11.27      (non-standard: accented cap height)
Comment TFM fontdimen 11: 1.94443    (non-standard: descender depth)
Comment TFM fontdimen 12: 11.27      (non-standard: max height)
Comment TFM fontdimen 13: 2.9        (non-standard: max depth)
Comment TFM fontdimen 14: 5          (non-standard: digit width)
Comment TFM fontdimen 15: 0.88889    (non-standard: uc stem)
Comment TFM fontdimen 16: 12         (non-standard: baselineskip)
Comment TFM fontdimen 17: 0.69444    (non-standard: lc stem)
Comment TFM fontdimen 18: 0.55556    (non-standard: u, i.e., font unit)
Comment TFM fontdimen 19: 0.22223    (non-standard: overshoot)
Comment TFM fontdimen 20: 0.25       (non-standard: thin stem, hair)
Comment TFM fontdimen 21: 0.30556    (non-standard: cap thin stem, i.e., cap_hair)
Comment TFM headerbyte  9: FontSpecific
Comment TFM headerbyte 49: LMRoman10
Comment TFM headerbyte 72: 234
\stopnarrowtyping

Watch the comments! Because \TEX\ needs a couple of so called fontdimens to be
set, the comments list the appropriate values. When a \type {tfm} file is
generated from an \type {afm} file, these values have to be used.

Each character (or glyph) gets an entry. When we run out of indices i.e.\ pass
the 255 boundary (we start at 0) the index becomes \type {-1}. Only the width is
specified. The height and depth have to be derived from the bounding box for
which the specification starts with key \type {B}.

\startnarrowtyping
StartCharMetrics 821
...
C  32 ; WX 333.33333 ; N space     ; B   0   0   0   0 ;
...
C 102 ; WX 305.55556 ; N f         ; B  33   0 357 705 ; L f ff ; L i fi ; L k f_k ; L l fl ;
C 105 ; WX 277.77777 ; N i         ; B  33   0 247 657 ;
C 108 ; WX 277.77777 ; N l         ; B  33   0 255 694 ;
...
C  -1 ; WX 500       ; N Acute     ; B 181 493 388 656 ;
C  -1 ; WX 500       ; N acute     ; B 188 510 374 698 ;
C  -1 ; WX 500       ; N acute.ts1 ; B 208 513 392 699 ;
...
EndCharMetrics
\stopnarrowtyping

Watch how this font defines a space character and keep in mind that these fonts
date from the time that there was only one kind of space. The \type {L} entry
specifies a ligature.

The names of glyphs are standardized, and even the \type {f_k} is conforming to
standards. This standardization makes it possible to go back from glyphs to characters
when copying text from a typeset document.

The kern table is pretty large here and for a reason. First of all the file
defines 821 glyphs so the average amount of kerns per glyph is not that large.
But take a look at the \type {A}. Because the \type {Aacute} has the same shape
it kerns in a similar way. This means that ideally all combined characters end up
with the same value as their base glyph. However, in our case a bit more
selective approach is taken. The \type {Adieresis} has a different set of kerns,
probably to save space. It is for this reason that \OPENTYPE\ fonts have a model
of kern classes so that similar shapes can be treated as one when setting kerns.
You see a similar issue with ligatures, where often the right part of the shape
kerns the same as the (stand alone) first part of the shape does.

\startnarrowtyping
StartKernData
StartKernPairs 9230
...
KPX seven.prop     hyphen.prop     -37
KPX seven.prop     four.prop       -74
KPX seven.prop     six.prop        -18.5
KPX hyphen.prop    one.prop        -37
KPX hyphen.prop    two.prop        -18.5
KPX hyphen.prop    seven.prop      -55.5
KPX seven.oldstyle four.oldstyle   -74
KPX A              T               -83.333
KPX Aacute         T               -83.333
KPX Abreve         T               -83.333
KPX Acircumflex    T               -83.333
...
KPX Adieresis      C               -27.778
...
KPX f              bracketright     27.778
KPX f              exclam           27.778
KPX f              parenright       27.778
KPX f              question         27.778
KPX f              quotedblleft     27.778
KPX f              quotedblleft.cm  27.778
KPX f              quotedblright    27.778
KPX f              quotedblright.cm 27.778
KPX f              quoteleft        27.778
KPX f              quoteright       27.778
...
KPX ff             bracketright     27.778
KPX ff             exclam           27.778
KPX ff             parenright       27.778
KPX ff             question         27.778
KPX ff             quotedblleft     27.778
KPX ff             quotedblleft.cm  27.778
KPX ff             quotedblright    27.778
KPX ff             quotedblright.cm 27.778
KPX ff             quoteleft        27.778
KPX ff             quoteright       27.778
...
EndKernPairs
EndKernData
\stopnarrowtyping

If you look closely at the names, you will notice that some glyphs have a
variant. In \OPENTYPE\ fonts these variants are grouped in features like \type
{oldstyle}. The first part of such a name is still part of the standardization,
but the second part is up to the font designer.

The file ends with:

\startnarrowtyping
EndFontMetrics
\stopnarrowtyping

\stopsection

\startsection[title=The \type {otf} file,reference=fontdata:otf]

In the \LUATEX\ manual you can find an overview of the raw \OTF\ format as
exposed in a \LUA\ table. The first decade of \LUATEX\ we used the built|-|in
loader but even then in \CONTEXT\ we didn't use that format directly but used it
to create a more compact and efficient table instead. The current release of
\CONTEXT\ uses its own loader written in \LUA, but the fundamentals have not
changed much. The tables are cached and can be read in at high speed. The
structure of the tables is unlikely to change much although more data might get
added. Although you can access the data it seldom makes sense to do so. Where
needed interfaces are provided.

\stopsection

\startsection[title=The \type {lfg} file,reference=fontdata:lua]

We use the goodies file control what gets added, replaced, patched or manipulated
in a font. A goodie file permits us to go beyond what font provide by default.
The content of a goodie file differs per font. As we also use this for
experiments, not all entries that you find in such files are meant for users.

\stopsection

\startsection[title=Used fonts]

The examples in the document depend on the fonts used. Here is a list of fonts used
to render this version. Because fonts might have changed in the meantime, some examples
might come out other than intended.

\startluacode
local list = table.load("fonts-mkiv-usedfonts.lua") or { }

local NC, NR = context.NC, context.NR

for i=1,#list do
    local l = list[i]
    context.starttabulate { "|lB|l|" }
        local basename  = string.lower(file.basename(l.filename))
        local version   = string.lower(l.version)
        local filesize  = (l.size or 0) / 1000 .. " Kb"
        local instances = l.instances or 1
        NC() context("filename")  NC() context.type(basename)  NC() NR()
        NC() context("instances") NC() context.type(instances) NC() NR()
        NC() context("filesize")  NC() context.type(filesize)  NC() NR()
        NC() context("version")   NC() context.type(version)   NC() NR()
    context.stoptabulate()
end
\stopluacode

\stopsection

\stopchapter

\stopcomponent

