% language=uk

\environment still-environment

\startcomponent still-simple

\startchapter[title=Removing something (typeset)]

\startsection[title=Introduction]

The primitive \type {\unskip} often comes in handy when you want to remove a
space (or more precisely: a glue item) but sometimes you want to remove more.
Consider for instance the case where a sentence is built up stepwise from data.
At some point you need to insert some punctuation but as you cannot look ahead it
needs to be delayed. Keeping track of accumulated content is no fun, and a quick
and dirty solution is to just inject it and remove it when needed. One way to
achieve this is to wrap this optional content in a box with special dimensions.
Just before the next snippet is injected we can look back for that box (that can
then be recognized by those special dimensions) and either remove it or unbox it
back into the stream.

To be honest, one seldom needs this feature. In fact I never needed it until
Alan Braslau and I were messing around with (indeed messy) bibliographic
rendering and we thought it would be handy to have a helper that could remove
punctuation. Think of situations like this:

\starttyping
John Foo, Mary Bar and others.
John Foo, Mary Bar, and others.
\stoptyping

One can imagine this list to be constructed programmatically, in which case the
comma before the \type {and} can be superfluous. So, the \type {and others} can
be done like this:

\startbuffer
\def\InjectOthers
  {\removeunwantedspaces
   \removepunctuation
   \space and others}

John Foo, Mary Bar, \InjectOthers.
\stopbuffer

\typebuffer

Notice that we first remove spaces. This will give:

\blank {\bf \getbuffer} \blank

where the commas after the names are coming from some not|-|too|-|clever automatism
or are the side effect of lazy programming. In the sections below I will describe
a bit more generic mechanism and also present a solution for non|-|\CONTEXT\ users.

\stopsection

\startsection[title=Marked content]

The example above can be rewritten in a more general way. We define a
couple macros (using \CONTEXT\ functionality):

\startbuffer
\def\InjectComma
  {\markcontent
     [punctuation]
     {\removeunwantedspaces,\space}}

\def\InjectOthers
  {\removemarkedcontent[punctuation]%
   \space and others}
\stopbuffer

\typebuffer \getbuffer

These can be used as:

\startbuffer
John Foo\InjectComma Mary Bar\InjectComma \InjectOthers.
\stopbuffer

\typebuffer

Which gives us:

\blank {\bf \getbuffer} \blank

Normally one doesn't need this kind of magic for lists because the length of the
list is known and injection can be done using the index in the list. Here is a more 
practical example:

\startbuffer
\def\SomeTitle {Just a title}
\def\SomeAuthor{Just an author}
\def\SomeYear  {2015}
\stopbuffer

\typebuffer \getbuffer

We paste the three snippets together:

\startbuffer
\SomeTitle,\space \SomeAuthor\space (\SomeYear).
\stopbuffer

\typebuffer \blank {\bf \getbuffer} \blank

But to get even more abstract, we can do this:

\startbuffer
\def\PlaceTitle
  {\SomeTitle
   \markcontent[punctuation]{.}}

\def\PlaceAuthor
  {\removemarkedcontent[punctuation]%
   \markcontent[punctuation]{,\space}%
   \SomeAuthor
   \markcontent[punctuation]{,\space}}

\def\PlaceYear
  {\removemarkedcontent[punctuation]%
   \space(\SomeYear)%
   \markcontent[punctuation]{.}}
\stopbuffer

\typebuffer \getbuffer

Used as:

\startbuffer
\PlaceTitle\PlaceAuthor\PlaceYear
\stopbuffer

\typebuffer

we get the output:

\blank {\bf \getbuffer} \blank

but when we have no author,

\startbuffer
\def\SomeAuthor{}

\PlaceTitle\PlaceAuthor\PlaceYear
\stopbuffer

\typebuffer

Now we get:

\blank {\bf \getbuffer} \blank

Even more clever is this:

\def\PlaceYear
  {\removemarkedcontent[punctuation]%
   \markcontent[punctuation]{\space(\SomeYear).}}

\startbuffer
\def\SomeAuthor{}
\def\SomeYear{}
\def\SomePeriod{\removemarkedcontent[punctuation].}

\PlaceTitle\PlaceAuthor\PlaceYear\SomePeriod
\stopbuffer

\typebuffer

The output is:

\blank {\bf \getbuffer} \blank

Of course we can just test for a variable like \type {\SomeAuthor} being empty
before we place punctuation, but there are cases where a period becomes a comma
or a comma becomes a semicolon. Especially with bibliographies your worst
typographical nightmares come true, so it is handy to have such a mechanism
available when it's needed.

\stopsection

\startsection[title=A plain solution]

For users of \LUATEX\ who don't want to use \CONTEXT\ I will now present an
alternative implementation. Of course more clever variants are possible but the
principle remains. The trick is simple enough to show here as an example of \LUA\
coding as it doesn't need much help from the infrastructure that the macro
package provides. The only pitfall is the used signal (attribute number) but you
can set another one if needed. We use the \type {gadgets} namespace to isolate
the code.

\startbuffer
\directlua {
  gadgets         = gadgets or { }
  local marking   = { }
  gadgets.marking = marking

  local marksignal   = 5001
  local lastmarked   = 0
  local marked       = { }
  local local_par    = 6
  local whatsit_node = 8

  function marking.setsignal(n)
    marksignal = tonumber(n) or marksignal
  end

  function marking.mark(str)
    local currentmarked = marked[str]
    if not currentmarked then
      lastmarked    = lastmarked + 1
      currentmarked = lastmarked
      marked[str]   = currentmarked
    end
    tex.setattribute(marksignal,currentmarked)
  end

  function marking.remove(str)
    local attr = marked[str]
    if not attr then
      return
    end
    local list = tex.nest[tex.nest.ptr]
    if list then
      local head = list.head
      local tail = list.tail
      local last = tail
      if last[marksignal] == attr then
        local first = last
        while true do
          local prev = first.prev
          if not prev or prev[marksignal] ~= attr or
               (prev.id == whatsit_node and
                  prev.subtype == local_par) then
             break
          else
            first = prev
          end
        end
        if first == head then
          list.head = nil
          list.tail = nil
        else
          local prev = first.prev
          list.tail  = prev
          prev.next  = nil
        end
        node.flush_list(first)
      end
    end
  end
}
\stopbuffer
\stopluacode

\typebuffer \getbuffer

These functions are called from macros. We use symbolic names for the marked
snippets. We could have used numbers but meaningful tags can be supported with
negligible overhead. The remover starts at the end of the current list and
goes backwards till no matching attribute value is seen. When a valid range is
found it gets removed.

\startbuffer
\def\setmarksignal#1%
  {\directlua{gadgets.marking.setsignal(\number#1)}}

\def\marksomething#1#2%
  {{\directlua{gadgets.marking.mark("#1")}{#2}}}

\def\unsomething#1%
  {\directlua{gadgets.marking.remove("#1")}}
\stopbuffer

\typebuffer \getbuffer

The working of these macros can best be shown from a few examples:

\startbuffer
before\marksomething{gone}{\em HERE}\unsomething{gone}after
before\marksomething{kept}{\em HERE}\unsomething{gone}after
\marksomething{gone}{\em HERE}\unsomething{gone}last
\marksomething{kept}{\em HERE}\unsomething{gone}last
\stopbuffer

\typebuffer

This renders as: \blank \startlines\bf\getbuffer\stoplines

The remover needs to look at the beginning of a paragraph marked by a local par
whatsit. If we removed that, \LUATEX\ would crash because the list head
(currently) cannot be set to nil. This is no big deal because this macro is not
meant to clean up across paragraphs.

A close look at the definition of \type {\marksomething} will reveal
an extra grouping in the definition. This is needed to make content that uses
\type {\aftergroup} trickery work correctly. Here is another example:

\startbuffer
\def\SnippetOne  {first\marksomething{punctuation}{, }}
\def\SnippetTwo  {second\marksomething{punctuation}{, }}
\def\SnippetThree{\unsomething{punctuation} and third.}
\stopbuffer

\typebuffer \getbuffer

We can paste these snippets together and make the last one use \type {and}
instead of a comma.

\startbuffer
\SnippetOne \SnippetTwo  \SnippetThree\par
\SnippetOne \SnippetThree\par
\stopbuffer

\typebuffer

We get: \blank {\bf \getbuffer} \blank

Of course in practice one probably knows how many snippets there are and using a
counter to keep track of the state is more efficient than first typesetting
something and removing it afterwards. But still it looks like a cool feature and
it can come in handy at some point, as with the title|-|author|-|year example given
before.

The plain code shown here is in the distribution in the file \type
{luatex-gadgets} and gets preloaded in the \type {luatex-plain} format.

\stopsection

\stopchapter
