% language=uk

\startcomponent cld-macros

\environment cld-environment

\startchapter[title=Files]

\startsection[title={Preprocessing}]

Although this option must be used with care, it is possible to preprocess files
before they enter \TEX. The following example shows this.

\starttyping
local function showline(str,filename,linenumber,noflines)
    logs.simple("[lc] file: %s, line: %s of %s, length: %s",
        file.basename(filename),linenumber,noflines,#str)
end

local function showfile(str,filename)
    logs.simple("[fc] file: %s, length: %s",
        file.basename(filename),#str)
end

resolvers.installinputlinehandler(showline)
resolvers.installinputfilehandler(showfile)
\stoptyping

Preprocessors like this are rather innocent. If you want to manipulate the
content you need to be aware of the fact that modules and such also pass your
code, and manipulating them can give unexpected side effects. So, the following
code will not make \CONTEXT\ happy.

\starttyping
local function foo()
    return "bar"
end

resolvers.installinputlinehandler(foo)
\stoptyping

But, as we pass the filename, you can base your preprocessing on names.

There can be multiple handlers active at the same time, and although more
detailed control is possible, the current interface does not provide that, simply
because having too many handlers active is asking for trouble anyway. What you
can do, is putting your handler in front or after the built in handlers.

\starttyping
resolvers.installinputlinehandler("before",showline)
resolvers.installinputfilehandler("after", showfile)
\stoptyping

Of course you can also preprocess files outside this mechanism, which in most
cases might be a better idea. However, the following example code is quite
efficient and robust.

\starttyping
local function MyHandler(str,filename)
    if file.suffix(filename) == "veryspecial" then
        logs.simple("preprocessing file '%s',filename)
        return MyConverter(str)
    else
        return str
    end
end

resolvers.installinputfilehandler("before",MyHandler)
\stoptyping

In this case only files that have a suffix \type {.veryspecial} will get an extra
treatment.

\stopsection

\stopchapter

\stopcomponent
