\usemodule[present-stepwise,present-bars,abr-01]

\startdocument
  [title=The scripts,
   color=darkyellow]

\StartSteps

\startsubject[title=Output]

    \startitemize[packed]

        \startitem \TEX\ collects content paragraph wise. \stopitem \FlushStep
        \startitem In between it can trigger the so called output routine. \stopitem \FlushStep
        \startitem At that moment you can do something with the result. \stopitem \FlushStep
        \startitem One of the things you can do is package all collected so far in a page. \stopitem \FlushStep

    \stopitemize

\stopsubject

\startsubject[title=Bonus]

    \startitemize[packed]

        \startitem In \LUATEX\ we can also intercept content at more places. \stopitem \FlushStep
        \startitem For instance before and after each paragraph is processed. \stopitem \FlushStep

    \stopitemize

\stopsubject

\StopSteps \page \StartSteps

\startsubject[title=Columns]

    \startitemize[packed]

        \startitem \TEX\ has no concept of columns. \stopitem \FlushStep
        \startitem You need to fake them by fiddling with the width and spitting boxes. \stopitem \FlushStep
        \startitem Often we can use tabulate (no output routine). \stopitem \FlushStep
        \startitem For some local applications we use simple columns. \stopitem \FlushStep
        \startitem In for instance itemize we used a mixed one- and multi-column model. \stopitem \FlushStep
        \startitem Columnsets are another (independent) mechanism, strongly grid based. \stopitem \FlushStep
        \startitem Traditional multicolumns are being replaced by a new mechanism: mixed columns. \stopitem \FlushStep

    \stopitemize

\stopsubject

\startsubject[title=Pitfalls]

    \startitemize[packed]

        \startitem Footnotes: page, first or last column, each column (delayed, immediate). \stopitem \FlushStep
        \startitem Graphics: moving floats around is more complex than in single columns. \stopitem \FlushStep
        \startitem Nesting: how about columns inside columns. \stopitem \FlushStep
        \startitem Balancing: can be hard taking all into account. \stopitem \FlushStep

    \stopitemize

\stopsubject

\StopSteps \page \StartSteps

\startsubject[title=Questions]

    \startitemize[packed]

        \startitem Do complex column mechanisms still make sense given the move to electronic paper. \stopitem \FlushStep
        \startitem If so, what functionality should be provided. \stopitem \FlushStep

    \stopitemize

\stopsubject

\StopSteps

\stopdocument
