\usemodule[present-stepwise,present-bars,abr-01]

\definecolor[verydark][s=.2]

\startdocument
  [title=Lexing sources,
   color=verydark]

\StartSteps

\startsubject[title=What is lexing]

    \startitemize[packed]
    \startitem Computer languages have mandate structure. \stopitem \FlushStep
    \startitem You can avoid errors by checking the input. \stopitem \FlushStep
    \startitem Editors can help by coloring reserved words, concept etc. \stopitem \FlushStep
    \startitem Consistency in coloring different languages makes sense. \stopitem \FlushStep
    \stopitemize

\stopsubject

\startsubject[title=When did we start]

    \startitemize[packed]
    \startitem We wrote our first editor begin 90's. \stopitem \FlushStep
    \startitem An extension quickly followed when we moved to \TEX: \TEXEDIT. \stopitem \FlushStep
    \startitem When \MODULA\ was no longer fashion we moved on to \PERL: \TEXWORK\ (quick demo) \stopitem \FlushStep
    \startitem When we ran into \SCITE\ we start using that. \stopitem \FlushStep
    \startitem I provided syntax highlighting for \TEX\ and \METAPOST\ (support for multiple formats etc.). \stopitem \FlushStep
    \stopitemize

\stopsubject

\StopSteps

\page

\StartSteps

\startsubject[title=Side effect of \MKIV]

    \startitemize[packed]
    \startitem \SCITE\ got \LPEG\ based lexing (external lexing). \stopitem \FlushStep
    \startitem I already had already written some lexers for the pretty printers. \stopitem \FlushStep
    \startitem So I gave it a go and made some more advanced lexers. \stopitem \FlushStep
    \startitem These ship with \CONTEXT: \TEX, \XML, \PDF, \LUA, \CLD, \METAPOST, text. \stopitem \FlushStep
    \stopitemize

\stopsubject

\startsubject[title=Characteristics]

    \startitemize[packed]
    \startitem The \TEX\ lexer supports nested lexing of \LUA\ and \METAPOST. \stopitem \FlushStep
    \startitem Integrated spell checking is provided. \stopitem \FlushStep
    \startitem Unfortunately there is no lexing for \SCITE\ on \MACOSX\ (not that I care too much nowadays). \stopitem \FlushStep
    \startitem It is a pitty that we have no access to \SCITE\ internal as with the regular \LUA\ interface. \stopitem \FlushStep
    \startitem On my good old machine huge files lex somewhat slow (at the end). \stopitem \FlushStep
    \stopitemize

\stopsubject

\StopSteps

\page

\StartSteps

\startsubject[title=The future]

    \startitemize[packed]
    \startitem I will improve the current lexers. \stopitem \FlushStep
    \startitem An \SQL\ lexer will be added at some point \stopitem \FlushStep
    \startitem I might make an \HTML/\CSS\ variant that supports nexted \LMX. \stopitem \FlushStep
    \stopitemize

\stopsubject

\StopSteps

\stopdocument
