% interface=en output=pdftex language=uk
%
% copyright=pragma-ade readme=readme.pdf licence=cc-by-nc-sa

\environment mcommon

\useurl[gpl-simple]   [http://creativecommons.org/licenses/GPL/2.0/]
\useurl[gpl-legal]    [http://creativecommons.org/licenses/GPL/2.0/legalcode]
\useurl[byncsa-simple][http://creativecommons.org/licenses/by-nc-sa/2.5/]
\useurl[byncsa-legal] [http://creativecommons.org/licenses/by-nc-sa/2.5/legalcode]

\useurl[garden]       [http://contextgarden.net]
\useurl[contextlist]  [http://www.ntg.nl/mailman/listinfo/ntg-context]
\useurl[development]  [http://www.ntg.nl/mailman/listinfo/dev-context]
\useurl[announce]     [http://www.ntg.nl/mailman/listinfo/ann-context]
\useurl[collector]    [http://context.literatesolutions.com]
\useurl[pragma]       [http://www.pragma-ade.com]
\useurl[mirror]       [http://context.aanhet.net]

\setupinteraction[state=start]

% copied from cont-log: readme_logo

\startuseMPgraphic{titlepage}{width,height}
  numeric width ; width = \MPvar{width} ;
  numeric height ; height = \MPvar{height} ;
  numeric delta ; delta := width/10 ;
  numeric circle ; circle := 2.5delta ;
  color c ; c := (.2,.4,.6) ;
  path p, q, r ;
  p := unitsquare xscaled width yscaled height ;
  z1 = (delta,height-2delta) ;
  z2 = (width-delta,height-delta) ;
  z3 = (width/2-delta,2delta+circle) ;
  z4 = (x3,delta+circle/2) ;
  q := z1 { dir  -15 } .. z2 &
       z2 { dir -105 } .. z3 &
       z3 { dir  135 } .. z1 &
       cycle ;
  r := fullcircle xscaled circle yscaled (.85circle) rotated 15 shifted z4 ;
  pickup pencircle scaled (delta/1.5) ;
  fill p withcolor .50c ;
  fill q withcolor .75c ;
  fill r withcolor .75c ;
  draw p withcolor    c ;
  draw q withcolor    c ;
  pickup pencircle scaled (delta/2) ;
  draw r withcolor    c ;
  setbounds currentpicture to p ;
\stopuseMPgraphic

\starttext

\TitlePage{titlepage}{90}{Read Me First}{}{1}

\subject {Introduction}

What licence suits best for a \TEX\ like system is a matter of
taste. Personally we dislike any licence that needs more than a few
pages of dense legal code to get the message accross. A \TEX\
related system like \CONTEXT\ is a hybrid of programs, scripts
and|/|or macro code as well as documentation and sample code,
including graphics. \TEX\ related systems also have a long
standing tradition of providing support structures for users. In
order to make support feasable, a \TEX\ based system like
\CONTEXT\ assumes a certain logic and structure in the way the
related files are named and organized in a tree structure. Even a
small change in one of the elements may let such a system behave
differently than manuals suggest. Swap a font, change some style
defaults, leave out some pieces, and users may end up in
confusion. A licence does not give a user any guarantees!

In order to satisfy those responsible for distributing \CONTEXT,
we need to choose a licence that makes them feel comfortable.
Unfortunately we don't feel that comfortable with a licence that does
not provide the guarantees that a system will not be adapted in
such ways that the advertised behaviour changes. On the
other hand, it is the responsibility of those distributing and
extending the system to make sure that this does not happen.
However, users should not automatically assume that what they get
shipped is the same as the original, which is why we stress that
support (from our side) will only be given on unaltered systems.

First of all, what is \CONTEXT ? It's just a bunch of macros,
written in \TEX\ and \METAPOST, meant for typesetting documents.
The macros are accompanied by some scripts, written in \PERL\ (mainly
the older scripts) \RUBY\ (the official ones) and \LUA\ (for
embedded usage). The \CONTEXT\ distribution comes with a few fonts,
files that help manage resources (e.g.\ map files), as well as
patterns (based on official ones, so this is a derived work).

The \CONTEXT\ distribution is packaged in a zip file organized in
the \TDS\ structure.

\starttabulate
\NC \type {cont-tmf.zip} \NC The main distribution. \NC \NR
\NC \type {cont-img.zip} \NC A few extra resources. \NC \NR
\NC \type {cont-ext.zip} \NC Third party modules.   \NC \NR
\stoptabulate

When we talk about \CONTEXT\ we also mean its graphical companion
\METAFUN\ and \FOXET, an \XML\ related product. All these are
included in the main distribution archive.

The documentation can be downloaded from our website, one of its
mirrors, the \TEX\ collection as distributed by \TEX\ user groups.
For some manuals, source code is available in a subversion
repository. The archives are also kept on \CTAN.

That said, what licence does apply? We need to distinguish between
things that resemble a program on the one hand and documentation
on the other hand. We (currently) use a different licence for
either of them.

\subject {The Code}

The program code (i.e. anything not under the \type {/doc}
subtree) is distributed under the

\startnarrower
\goto{Creative Commons GNU GPL}[url(gpl-simple)]
\stopnarrower

For practical purposes distributers may also choose the \LATEX\
project licence, which is considered to be a bit more \TEX\
friendly. (BSD alike licences, the Ruby Licence and the Apache
are all licences that apply well for \CONTEXT.)

In practice, users may forget about the legal part, if only
because I haven't even read (and understood) it completely myself,
so let's stick to what Creative Commons makes of it:

\startcolor[blue]
The GNU General Public License is a Free Software license. Like
any Free Software license, it grants to you the four following
freedoms:

\startitemize
\item The freedom to run the program for any purpose.
\item The freedom to study how the program works and adapt it to
      your needs.
\item The freedom to redistribute copies so you can help your neighbour.
\item The freedom to improve the program and release your improvements
      to the public, so that the whole community benefits.
\stopitemize

You may exercise the freedoms specified here provided that you
comply with the express conditions of this license. The principal
conditions are:

You must conspicuously and appropriately publish on each copy
distributed an appropriate copyright notice and disclaimer of
warranty and keep intact all the notices that refer to this
License and to the absence of any warranty; and give any other
recipients of the Program a copy of the GNU General Public License
along with the Program. Any translation of the GNU General Public
License must be accompanied by the GNU General Public License.

If you modify your copy or copies of the program or any portion of
it, or develop a program based upon it, you may distribute the
resulting work provided you do so under the GNU General Public
License. Any translation of the GNU General Public License must be
accompanied by the GNU General Public License.

If you copy or distribute the program, you must accompany it with
the complete corresponding machine-readable source code or with a
written offer, valid for at least three years, to furnish the
complete corresponding machine-readable source code.

Any of these conditions can be waived if you get permission from
the copyright holder.

Your fair use and other rights are in no way affected by the above.
\stopcolor

\subject {Recommendations}

Here are a few recommendations in case you want to distribute,
extend of embed \CONTEXT\ in applications:

\startitemize

\item You can best leave the code base untouched. Most of
\CONTEXT\ provides hooks and it's relatively easy to overload
code. Leave the lower level system code untouched: changes may
backfire when you update. Asking for more hooks is the best way to
go.

\item Put your own code in the right subpaths, i.e.\ modules
approved by the development team under \type {.../third}, and
styles and whatever else under \type {.../user}. This way your
code will not interfere with existing code and updating will give
less problems. Keep in mind that \TEX\ systems have their own way
and order in locating files, and the load order often matters.

\item Don't copy styles and change a few lines, but load the base one
and built|/|patch on top of that. In the end you may benefit from
improvements to the base style.

\item Be original. The whole idea behind \CONTEXT\ is that you can
write your own styles. On the \CONTEXT\ mailing list as well as on
the Wiki there are enough advanced users to help you make a start.

\item Don't hesitate to submit bugs reports and ask for
extensions. It may even be that what you want is already present
but yet undocumented.

\item If things don't work as expected, check to what extend your
system matches the (more or less) standard. We provide so called
minimal \CONTEXT\ trees that can serve as a reference. Because
\CONTEXT\ evolves, make sure your system is up to date.

\item The scripts can best be called using \type {texmfstart}. This
lessens dependencies on the location in the tree and ensures upward
compatibility. It also prevents clashes with similary named scripts.

\item Some scripts depend on each other. Don't mess around with the
existing functionality and names of the scripts and then feed them
back into the standard distributions.

\stopitemize

\subject {Documents}

The documentation is provided under another Creative Commons licence

\startnarrower
\goto{Attribution NonCommercial ShareAlike}[url(byncsa-simple)]
\stopnarrower

This one says:

\startcolor[blue]
You are free:

\startitemize
\item to copy, distribute, display, and perform the work
\item to make derivative works
\stopitemize

{\sc Attribution:} You must attribute the work in the manner
specified by the author or licensor.

{\sc NonCommercial:} You may not use this work for commercial
purposes.

{\sc Share Alike:} If you alter, transform, or build upon this
work, you may distribute the resulting work only under a license
identical to this one.

\startitemize
\item For any reuse or distribution, you must make clear to others
      the license terms of this work.
\item Any of these conditions can be waived if you get permission from
      the copyright holder.
\stopitemize

Your fair use and other rights are in no way affected by the above.
\stopcolor

The non||commercial part is mostly a safeguard. We don't mind if
user groups distribute printed copies, publish (parts of) manuals
and|/|or if authors use example code in manuals and books about
\CONTEXT.

If you distribute \CONTEXT\ and related software on electronic media
as part of \TEX\ distributions (either or not for money), you may
also distribute the manuals and their sources in electronic form,
preferable as provided by the maintainers of \CONTEXT.

Keep in mind that logos and cover designs are not meant to be
copied. We provide the source code for some manuals, but we don't
always provide all graphics and other resources. For instance, in
some manuals we use commercial fonts and you have to buy those
yourself.

We provide the typeset manuals at our website. Those are the official
ones. We appreciate it if you do not to distribute manuals compiled
on your own system as substitutes. The manuals are a showcase for what
\CONTEXT\ provides. Help us to assure the quality.

\subject {More information}

We're not going to fill \mathematics{n}~pages with legal stuff, so if
you want to know more, you have to consult the web for the legalities
mentioned. Here are a few starting points:

\startlines
\goto{\url[gpl-simple]}[url(gpl-simple)]
\goto{\url[gpl-legal]}[url(gpl-legal)]
\stoplines

\startlines
\goto{\url[byncsa-simple]}[url(byncsa-simple)]
\goto{\url[byncsa-legal]}[url(byncsa-legal)]
\stoplines

\CONTEXT\ itself can be fetched from the main site or its primary mirror:

\startlines
\goto{\url[pragma]}[url(pragma)]
\goto{\url[mirror]}[url(mirror)]
\stoplines

A starting point for support can be found at:

\startlines
\goto{\url[contextlist]}[url(contextlist)]
\goto{\url[garden]}[url(garden)]
\stoplines

Bugs and feature requests can be registered at the collector:

\startlines
\goto{\url[collector]}[url(collector)]
\stoplines

Releases are announced at:

\startlines
\goto{\url[announce]}[url(announce)]
\stoplines

The developers can be met at:

\startlines
\goto{\url[development]}[url(development)]
\stoplines

\subject {Disclaimer}

To play safe we include a disclaimer here, taken from the BSD style
licence. For some reason such a text is in capitals, so \unknown

\start \sc \blue
THIS SOFTWARE IS PROVIDED BY THE AUTHOR \quotation {AS IS} AND ANY EXPRESS OR
IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES
OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED.
IN NO EVENT SHALL THE AUTHOR BE LIABLE FOR ANY DIRECT, INDIRECT,
INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
(INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
\stop

\unknown\ and don't bother discussing licence issues and related things
with us for the mere sake of discussing licence stuff.

\blank[2*big]

\startlines
Hans Hagen
PRAGMA ADE
Hasselt NL
\stoplines

\ColofonPage

\stoptext
