% language=uk

\environment details-environment

\startcomponent details-colofon

\startchapter[title={About this document}]

This document is typeset in \CONTEXT\ using \LUATEX\ with \METAPOST. We use only
one font: the Computer Modern Typewriter. The verbatim portions of the text are
typeset in its mono spaced variant. One of the reasons that I chose this font is
that we need a mono spaced font to typeset the example code, and the Computer
Modern Typewriter is one the best there is. This font combines well with many
other typefaces, but the sometimes excessive use of different fonts (and sizes)
in the styles that I have to implement made me long for simplicity. And so I
decided to stick to one font. A careful reader will notice that this document has
character protruding enabled (resulting in hanging punctuation).

We use a couple of colors. Again, I went for simplicity and use rather primary
colors, although I do use them in transparent variants as well.

There is not much more to say, apart from that I want to thank our customers as
well as \CONTEXT\ users for asking me to implement \DTP\ competing styles and
features. Their demands drive \CONTEXT\ in directions we could not have foreseen
when we started its development.

We use a (transparent) gray background behind the text so that we have an
indication where the text area is positioned relative to the page. It also
enables us to comfortably turn on the grid.

Some features shown here are relatively new and therefore they occasionally are
improved. As a result some aspects of their functionality may change.

\stopchapter

\stopcomponent
