% language=uk

\startcomponent languages-options

\environment languages-environment

\startchapter[title=Options][color=darkblue]

\startsection[title=Introduction]

Hyphenation of words is controlled by so called patterns. They take a word and
try to match parts with a pattern that describes where a hyphen can be injected.
Preferred and discouraged injection points accumulate to a score that in the end
determine where so called discretionary nodes gets injected in the list of
glyphs that make a word. The patterns are language specific.

This mechanism is agnostic when it comes to the characters involved: they are
just numbers. However, when in a next step font features like ligature building
and kerning are applied we also have to deal with language specific properties
(and meanings). Often a ligature at the boundary of a composed word can make
reading confusing and has to be avoided. Some of that can be controlled by the
font when it implements language specific features but because that approach is
not based on a dictionary it is more about playing safe and prevention than about
quality.

In the next sections a mechanism is discussed that also uses patterns. This time
it is about controlling fonts as well as how hyphenation patterns are applied.
This process kicks in before hyphenation is applied but it definitely has to be
seen as part of that same process. It is integrated in hyphenation machinery and
acts as preprocessor with the possibility to feedback and move forward. The
implementation is such that when it's not used there is no performance penalty.
\footnote {There are by now plenty of alternative approaches to these problems
but after some discussion about the pro's and cons of each this new mechanism was
made. I admit that the fun factor played a role. It is also one of the things we
can do in \LUAMETATEX\ without worrying about a possible negative impact on
\LUATEX\ users other than \CONTEXT .}

There are several predefined operations that are characterized by keywords and
shortcuts and collected in an option list that is part of a language goodie file.
Examples can be found in the distribution in files with the suffix \type {llg}
(\LUA\ language goodie). The framework of such a file is:

\starttyping
return {
    name       = "whatever",
    version    = "1.00",
    comment    = "Goodies for experiments and demo.",
    author     = "Hans Hagen",
    copyright  = "ConTeXt development team",
    options    = {
        { ... },
        ........
        { ... },
    }
}
\stoptyping

These options will eventually result in patterns that are bound to words,
think of:

\starttabulate[|T||||]
\NC effe     \NC \type {foo|bar}   \NC \type {..|..}     \NC inhibit ligature \NC \NR
\NC foobar   \NC \type {foo=bar}   \NC \type {...=...}   \NC inhibit kerning  \NC \NR
\NC somemore \NC \type {some+more} \NC \type {....+....} \NC compound word    \NC \NR
\stoptabulate

The whole repertoire is:

\starttabulate[||T|]
\NC \type {a|b} \NC a:norightligature, b:noleftligature \NC \NR
\NC \type {a=b} \NC a:norightkern, b:noleftkern         \NC \NR
\NC \type {a<b} \NC b:noleftkern                        \NC \NR
\NC \type {a>b} \NC a:norightkern                       \NC \NR
\NC \type {a+b} \NC a:compound:b                        \NC \NR
\stoptabulate

Later we will see how some can be combined. An option can be defined using entries
in a subtable:

\starttabulate[|T|||]
\NC patterns   \NC hash            \NC \type {[snippet] = "replacement pattern"} \NC \NR
\NC words      \NC string          \NC string of words, separated by whitespace \NC \NR
\NC prefixes   \NC string          \NC snippets that combine with words (at the start) \NC \NR
\NC suffixes   \NC string          \NC snippets that combine with words (at the end) \NC \NR
\NC matches    \NC array or number \NC a number or table indicating which match matters \NC \NR
\NC actions    \NC hash            \NC \type {[character] = "action(s)"} \NC \NR
\NC characters \NC string          \NC permitted characters (additional hjcodes) \NC \NR
\NC return     \NC integer         \NC what to do next \NC \NR
\stoptabulate

The default return value is~2 but there are some more:

\starttabulate[|T||]
\NC 0 \NC go to the next (valid) word \NC \NR
\NC 1 \NC restart \NC \NR
\NC 2 \NC exceptions and after that patterns \NC \NR
\NC 3 \NC patterns \NC \NR
\stoptabulate

There are some safeguards built in that force a restart. For instance when a word
is replaced a restart is enforces unless we skip the word. A restart will not
permit a second replacement (after all we need to avoid endless loops).

In a multi|-|line word list, lines that start with a comment trigger: \LUA's
double dash or the usual \TEX\ percent sign.

\stopsection

\startsection[title=Inhibiting]

The next definition replaces \type {ff} by \type {f|f} in the words given and
eventually block a ligature.

\starttyping
{
    patterns = {
        ff  = "f|f",
    },
    words = [[
        effe
    ]],
}
\stoptyping

Some fonts provide the \type {ij} ligature or do some special kerning between
these characters (something Dutch). Because it depends on the font logic if a
dedicated replacement or kerning is used this is an example where we do this:

\starttyping
{
    patterns = {
        ij = "i|j",
    },
    actions = {
        ["|"] = "nokern noligature",
    },
    words = [[
        ijverig
     -- fijn -- to ligature fi or ij, that's the question
    ]],
}
\stoptyping

A more extensive definition is the following. Here we explicitly define that only
the first match in a word get treated. Here we not only block ligatures but also
kerns.

\starttyping
{
    patterns = {
        ff  = "f|f",
    },
    matches = { 1 },
    actions = {
        ["|"] = "noligature nokern"
    },
    words = [[
        effe
        effeffe
    ]],
}
\stoptyping

You can also omit the pattern when you inject specifiers yourself:

\starttyping
{
    actions = {
        ["|"] = "noligature nokern"
    },
    words = [[
        ef|fe
        ef|fef|fe
    ]],
}
\stoptyping

You can also use different shortcuts:

\starttyping
{
    actions = {
        ["1"] = "noligature"
        ["2"] = "nokern"
    },
    words = [[
        ef1fe
        ef1fef2fe
    ]],
}
\stoptyping

Although I cannot come up with a nice example, there can be reasons for
inhibiting kerns. Here we inhibit kerns left of the upcoming character:

\starttyping
{
    patterns = {
        fo = "f<o",
        rm = "r<m",
    },
    words = [[
        information
    ]],
}
\stoptyping

And here we inhibit kerns left of the previous and upcoming character:

\starttyping
{
    patterns = {
        th = "t=h",
    },
    words = [[
        thrive
    ]],
}
\stoptyping

Just look in the files in the distribution for realistic examples, like

\starttyping
{
    patterns = {
        fi = "f|i",
    },
    words = [[
        deafish dwarfish elfish oafish selfish
    ]],
    suffixes = [[
        ness ly
    ]]
}
\stoptyping

where we block ligatures in 15 words. There's also a \type {prefixes} key.

\stopsection

\startsection[title=Replacements]

Replacements are probably not used that much but here is one for German. Not
only is the uppercase variant of ß seldom used, many fonts don't provide it
so we can best replace it:

\starttyping
{
    characters = "ẞ", -- uppercase ß, not visible in all verbatim fonts
    patterns   = {
        ["ẞ"] = "SS", -- key is uppercase ß
    },
}
\stoptyping

Here we define that character as valid, something that normally is done with the
patterns but patterns don't have them. If we do not specify it here, the
hyphenator will skip this word. For the record: this can also be done with a font
feature that decomposes the character.

\stopsection

\startsection[title=Compound words]

You might want to suppress ligatures and maybe even kerning when compound words
are involved.

\starttyping
{
    patterns = {
        ff = "f+f",
    },
    words = [[
        aaaaffaaaa
        bbffbb
    ]],
}
\stoptyping

Again you can also say:

\starttyping
{
    words = [[
        aaaaf|faaaa
        bbf|fbb
    ]],
}
\stoptyping

But patterns make sense when you have a large list (that might come from some
other source than yourself).

The next specification will turn two times three \type {bla}'s into a compound
word but also make sure that we have at least 4 characters left and right of a
potential break.

\starttyping
    {
        left  = 4,
        right = 4,
        words = [[
            blablabla+blablabla
        ]],
    }
\stoptyping

\stopsection

\startsection[title=Performance]

Although these mechanisms introduce overhead, the performance hit in \LMTX\ is
not that large. This is because the number of words in a document is limited and
\LUA\ is fast enough.

\stopsection

\startsection[title=Plugins]

{\em This interface is preliminary but for the record I put an example here
anyway.}

\starttyping
local n = 0
function document.myhack(original)
    n = n + 1
    print(n,original)
    return original
end

languages.installhandler("de","document.myhack")
\stoptyping

One can manipulate a text as in:

\starttyping
function document.myhack(original)
    local t = utf.split(original)
    local t = table.reverse(t)
    local f = t[#t]
    local l = t[1]
    if characters.upper(f) == f then
        t[1]  = characters.upper()
        t[#t] = characters.lower(f)
    end
    local original = table.concat(t)
    return original
end

languages.installhandler("en","document.myhack")
\stoptyping

The text will fed again into the hyphenator and treated in the normal way. There
are some safeguards against the text being processed twice.

\stopsection

\startsection[title=Tracing]

You can also embed definitions in the source file:

\starttyping
\startlanguageoptions[de]
    Zapf|innovation
\stoplanguageoptions
\stoptyping

\stopsection

\startsection[title=Exceptions]

When you set exceptions in a goodie file, it will use the plugin mechanism to
check for them. This is a bit more efficient than using the internal checkerm
which actually also goes via a\LUA\ hash.

\starttyping
{
    exceptions = [[
        a-very{-}{-}{w}eird{1}{2}{3}(w)ord
    ]],
}
\stoptyping

Watch out: when you specify a discretionary replacement three braced valued are
passed: the pre, post and replace text. The replace text is used in the lookup,
unless you add a string between parentheses, which then will be used instead. A
digit between bracket will apply a penalty according to the following logic (in
the engine): A zero digit results in \type {\hyphenpenalty}, otherwise the
digits~1 upto~9 will be used as multiplier for \type {\exceptionpenalty} when
that value is larger than 100000, otherwise \type {\exceptionpenalty} is used.

\stopsection

\startsection[title=Tracing]

The following tracker can be used:

\starttyping
\enabletrackers[languages.goodies]
\stoptyping

In addition the style \type {languages-goodies} implements some tracing options.
You can just run that one to see what it does.

The engine itself has also a tracing option: \type {\tracinghyphenation}. When
set to zero nothing is shown, when set to one redundant patterns will be
reported. A value of two reports what words get fed into the hyphenator and if
they got hyphenated. A value of three gives more detail: when a word gets
hyphenated the relevant (resulting) part of the node list is shown. You need to
set \type {\tracingonline} to a value larger than zero to get this reported to
the console. Expects lots of extra output to the console for large documents but
it can be revealing.

\stopsection

\stopchapter

\stopcomponent

%D Musical timestamp: end Match 2021: running into Joe Parrish's amazing
%D interpretation of Stravinsky's "Rite of Spring" on guitars.
%D
%D Also on YT: The Rite of Spring by London Symphony Orchestra (conducted
%D by Simon Rattle).
