% language=uk

\startcomponent musings-whytex

\environment musings-style

\startchapter[title={Why use \TEX ?}]

\startsection[title={Introduction}]

Let's assume that you know what \TEX\ is: a program that interprets a language
with the same name that makes it possible to convert (tagged) input into for
instance \PDF. For many of its users it is a black box: you key in some text, hit
a button and get some typeset result in return. After a while you start tweaking
this black box, meet other users (on the web), become more fluent and stick to it
forever.

But now let's assume that you don't know \TEX\ and are in search of a system
that helps you create beautiful documents in an efficient way. When your
documents have a complex structure you are probably willing to spend some time on
figuring out what the best tool is. Even if a search lets you end up with
something called \TEX, a three letter word with a dropped E, you still don't
know what it is. Advertisement for \TEX\ is often pretty weak. It's rather easy
to point to the numerous documents that can be found on the web. But what exactly
does \TEX\ do and what are its benefits? In order to answer this we need to know
who you are: an author, editor, an organization that deals with documents or needs
to generate readable output, like publishers do.

\stopsection

\startsection[title={Authors}]

We start with authors. Students of sciences that use mathematics don't have much
of a choice. But most of these documents hardly communicate the message that
\quotation {Everyone should use \TEX.} or that \quotation {All documents produced
by \TEX\ look great.} but they do advocate that for rendering math it is a pretty
good system. The source code of these documents often look rather messy and
unattractive and for a non|-|math user it can be intimidating. Choosing some
lightweight click|-|and|-|ping alternative looks attractive.
lightweight click|-|and|-|ping alternative looks attractive.

Making \TEX\ popular is not going to happen by convincing those who have to write
an occasional letter or report. They should just use whatever suits them. On the
other hand if you love consistency, long term support, need math, are dealing
with a rare language or script, like to reuse content, prefer different styling
from one source, use one source for multiple documents, or maybe love open source
tools, then you are a candidate. Of course there is a learning curve but normally
you can master \TEX\ rather fast and once you get the hang of it there's often no
way back. But you always need to invest a bit beforehand.

So what authors are candidates for \TEX ? It could be that \TEX\ is the only tool
that does the job. If so, you probably learned that from someone who saw you
struggle or had the same experience and wrote or talked about it somewhere. In
that case using \TEX\ for creating just one document (like a thesis) makes sense.
Otherwise, you should really wonder if you want to invest time in a tool that you
probably have to ditch later on as most organizations stick to standard
(commercial) word processing tools.

Talking to customers we are often surprised that people have heard about \TEX, or
even used it for a few documents in college. Some universities just prescribe the
use of \TEX\ for reporting, so not much of a choice there. Memories are normally
rather positive in the sense that they know that it can do the job and that it's
flexible.

User group journals, presentations at \TEX\ meetings, journals, books and manuals
that come with \TEX\ macro packages can all be used to determine if this tool
suits an author. Actually, I started using \TEX\ because the original \TEX book
had some magic, and reading it was just that: reading it, as I had no running
implementation. A few years later, when I had to write (evolving) reports, I
picked up again. But I'm not a typical user.

\stopsection

\startsection[title={Programmers}]

When you are a programmer who has to generate reports, for instance in \PDF, or
write manuals, then \TEX\ can really be beneficial. Of course \TEX\ is not always
an obvious choice, but if you're a bit able to use it it's hard to beat in
quality, flexibility and efficiency. I'm often surprised that companies are
willing to pay a fortune for functionality that basically comes for free.
Programmers are accustomed to running commands and working in a code editor with
syntax highlighting so that helps too. They too recognize when something can be
done more efficiently.

When you need to go from some kind of input (document source, database,
generated) to some rendered output there currently are a few endpoints: a
(dynamic) \HTML\ page, a \PDF\ document, something useable in a word processor,
or a representation using the desktop user interface. It's the second category
where \TEX\ is hard to beat but even using \TEX\ and \METAPOST\ for creating a
chart can make sense.

There are of course special cases where \TEX\ fits in nicely. Say that you have
to combine \PDF\ documents. There are numerous tools to do that and \TEX\ is one.
The advantage of \TEX\ over other tools is that it's trivial to add additional
text, number pages, provide headers and footers. And it will work forever. Why?
Because \TEX\ has been around for decades and will be around for decades to come.
It's an independent component. The problem with choosing for \TEX\ is that the
starting point is important. The question is not \quotation {What tool should I
use?} but \quotation {What problem do I need to solve?}. An open discussion about
the objectives and possibilities is needed, not some checklist based on
assumptions. If you don't know \TEX\ and have never worked with a programmable
typesetting environment, you probably don't see the possibilities. In fact, you
might even choose for \TEX\ for the wrong reasons.

The problem with this category of users is that they seldom have the freedom to
choose their tools. There are not that many jobs where the management is able to
recognize the clever programmer who can determine that \TEX\ is suitable for a
lot of jobs and can save money and time. Even the long term availability and
support is not an argument since not only most tools (or even apis) changes every
few years but also organizations themselves change ownership, objectives, and
personnel on a whim. The concept of \quote {long term} is hard to grasp for most
people (just look at politics) and it's only in retrospect that one can say
\quote {We used that toolkit for over a decade.}

\stopsection

\startsection[title={Organizations}]

Authors (often) have the advantage that they can choose themselves: they can use
what they like. In practice any decent programmer is able to find the suitable
tools but convincing the management to use one of them can be a challenge. Here
we're also talking of \quote {comfort zones}: you have to like a tool(chain).
Organizations normally don't look for \TEX. Special departments are responsible
for choosing and negotiating whatever is used in a company. Unfortunately
companies don't always start from the open question \quotation {We have this
problem, we want to go there, what should we do?} and then discuss options with
for instance those who know \TEX. Instead requirements are formulated and matches
are found. The question then is \quotation {Are these requirements cut in stone?}
and if not (read: we just omit some requirements when most alternatives don't
meet them), were other requirements forgotten? Therefore organizations can end up
with the wrong choice (using \TEX\ in a situation where it makes no sense) or
don't see opportunities (not using \TEX\ while it makes most sense). It doesn't
help that a hybrid solution (use a mix of \TEX\ and other tools) is often not an
option. Where an author can just stop using a tool after a few days of
disappointment, and where a programmer can play around a bit before making a
choice, an organization probably best can start small with a proof of concept.

Let's take a use case. A publisher wants to automatically convert \XML\ files
into \PDF. One product can come from multiple sources (we have cases where
thousands of small \XML\ files combine into one final product). Say that we have
three different layouts: a theory book, a teachers manual and an answer book. In
addition special proofing documents have to be rendered. The products might be
produced on demand with different topics in any combination. There is at least
one image and table per page, but there can be more. There are color and
backgrounds used, tables of contents generated, there is extensive cross
referencing and an index. Of course there is math.

Now let's assume an initial setup costs 20K Euro and, what happens often when the
real products show up, a revision after one year takes the same amount. We also
assume 10K for the following eight years for support. So, we end up with 120K
over 10 years. If one goes cheap we can consider half of that, or we can be
pessimistic and double the amount.

The first year 10K pages are produced, the second year 20K and after that 30K per
year. So, we're talking of 270K pages. If we include customer specific documents
and proofing we might as well end up with a multiple of that.

So, we have 120K Euro divided by 270K pages or about half an Euro per page. But
likely we have more pages so it costs less. If we double the costs then we can
assume that some major changes took place which means more pages. In fact we had
projects where the layout changed, all documents were regenerated and the costs
were included in the revision, so far from double. We also see many more pages
being generated so in practice the price per page drops below half an Euro. The
more we process the cheaper it gets and one server can produce a lot of pages!

Now, the interesting bit of such a calculation is that the costs only concern the
hours spent on a solution. A \TEX\ based system comes for free and there are no
license costs. Whatever alternative is taken, even if it is as flexible, it will
involve additional costs. From the perspective of costs it's very hard to beat
\TEX. Add to that the possibility for custom extensions, long term usage and the
fact that one can adapt the system. The main question of course is: does it do
the job. The only way to find out is to either experiment (which is free),
consult an expert (not free, but then needed anyway for any solution) or ask an
expert to make a proof of concept (also not free but relatively cheap and
definitely cheaper than a failure). In fact, before making decisions about what
solution is best it might be a good idea to check with an expert anyway, because
more or less than one thinks might be possible. Also, take into account that the
\TEX\ ecosystem is often one of the first to support new technologies, and
normally does that within its existing interface. And there is plenty of free
support and knowledge available once you know how to find it. Instead of wasting
time and money on advertisement and fancy websites, effort goes into support and
development. Even if you doubt that the current provider is around in the decade
to come, you can be sure that there will be others, simply because \TEX\ attracts
people. Okay, it doesn't help that large companies like to out source to
far||far||away and expect support around the corner, so in the end they might
kill their support chain.

When talking of \TEX\ used in organizations we tend to think of publishers. But
this is only a small subset of organizations where information gets transformed
into something presentable. For small organizations the choice for \TEX\ can be
easy: costs, long term stability, knowing some experts are driving forces. For
large organizations these factors seem (at least to us) hardly relevant. We've
(had) projects where actually the choice for using a \TEX\ based solution was (in
retrospect) a negative one: there was no other tool than this relatively unknown
thing called \TEX. Or, because the normal tools could not be used, one ended up
with a solution where (behind the scenes) \TEX\ is used, without the organization
knowing it. Or, it happened that the problem at hand was mostly one that demands
in|-|depth knowledge of manipulating content, cleaning up messy data, combining
resources (images or \PDF\ documents), all things that happen to be available in
the perspective of \TEX. If you can solve a hard to solve problem for them then
an organization doesn't care what tool you use. What does matter is that the
solution runs forever, that costs are controllable and above all, that it
\quotation {Just works.} And if you can make it work fast, that helps too. We
can safely claim that when \TEX\ is evaluated as being a good option, that in the
end it always works out quite well.

Among arguments that (large) organizations like to use against a choice for \TEX\
(or something comparable) are the size of the company that they buy their
solution from, the expected availability for support, and the wide|-|spread usage
of the tool at hand. One can wonder if it also matters that many vendors change
ownership, change products every few years, change license conditions when they
like, charge a lot for support or just abort a tool chain. Unfortunately when that
happens those responsible for choosing such a system can have moved on to another
job, so this is seldom part of an evaluation. For the supplier the other side of
the table is just as much of a gamble. In that respect, an organization that
wants to use an open source (and|/|or free) solution should realize that getting
a return on investment on such a development is pretty hard to achieve. So, who
really takes the risk for writing open source?

For us, the reason to develop \CONTEXT\ and make it open is that it fits in our
philosophy and we like the community. It is actually not really giving us an
advantage commercially: it costs way more to develop, support and keep
up|-|to|-|date than it will ever return. We can come up with better, faster and
easier solutions and in the end we pay the price because it takes less time to
cook up styles. So there is some backslash involved because commercially a
difficult solution leads to more billable hours. Luckily we tend to avoid wasting
time so we improve when possible and then it ends up in the distributed code.
And, once the solution is there, anyone can use it. Basically also for us it's
just a tool, like the operating system, editor and viewer are. So, what keep
development going is mostly the interaction with the community. This also means
that a customer can't really demand functionality for free: either wait for it to
show up or pay for it (which seldom happens). Open source is not equivalent with
\quotation {You get immediately what you want because someone out there writes
the code.}. There has to be a valid reason and often it's just users and meetings
or just some challenge that drives it.

This being said, it is hard to convince a company to use \TEX. It has to come
from users in the organization. Or, what we sometimes see with publishers, it
comes with an author team or acquired product line where it's the only option.
Even then we seldom see transfer to other branches in the organizations. No one
seems to wonder \quotation {How on earth can that \XML\ to \PDF\ project produce
whatever output in large quantities in a short period of time} while other (past)
projects failed. It probably relates to the abstraction of the process. Even
among \TEX\ users it can be that you demonstrate something with a click on a
button and that many years afterwards someone present at that moment tells you
that they just discovered that this or that can be done by hitting a button. I'm
not claiming that \TEX\ is the magic wand for everything but in some areas it's
pretty much ahead of the pack. Go to a \TEX\ user meeting and you will be surprised
about the accumulated diverse knowledge present in the room. It's user demand that
drives \CONTEXT\ development, not commerce.

\stopsection

\startsection[title={Choosing}]

So, where can one find information about \TEX\ and friends? On the web
one has to use the right search keys, so adding \type {tex} helps: \typ {context
tex} or \typ {xml tex pdf} and so on. Can one make a fancy hip website, sure, but
it being a life|-|long, already old and mature environment, and given that it
comes for free, or is used low|-|budget, not much effort and money can be spent
on advertising it. A benefit is that no false promises and hypes are made either.
If you want to know more, just ask the right folks.

For all kind of topics one can find interesting videos and blogs. One can
subscribe to channels on YouTube or join forums. Unfortunately not that many
bloggers or vloggers or podcasters come up with original material every time, and
often one starts to recognize patterns and will get boring by repetition of wisdom
and arguments. The same is true for manuals. Is a ten year old manual really
obsolete? Should we just recompile it to fake an update while in fact there has
been no need for it? Should we post twenty similar presentations while one can
do? (If one already wants to present the same topic twenty times in the first
place?) Maybe one should compare \TEX\ with cars: they became better over time
and can last for decades. And no new user manual is needed.

As with blogs and vlogs advertising \TEX\ carries the danger for triggering
political discussions and drawing people into discussions that are not pleasant:
\TEX\ versus some word processor, open versus closed source, free versus paid
software, this versus that operating system, editor such or editor so.

To summarize, it's not that trivial to come up with interesting information about
\TEX, unless one goes into details that are beyond the average user. And those
who are involved are often involved for a long time so it gets more complex over
time. User group journals that started with tutorials later on became expert
platforms. This is a side effect of being an old and long|-|term toolkit. If
you run into it, and wonder if it can serve your purpose, just ask an expert.

Most \TEX\ solutions are open source and come for free as well. Of course if you
want a specific solution or want support beyond what is offered on mailing lists
and forums you should be willing to pay for the hours spent. For a professional
publisher (of whatever kind) this is not a problem, if only because any other
solution also will cost something. It is hard to come up with a general estimate.
A popular measure of typesetting costs is the price per page, which can range
from a couple of euro's per page to two digit numbers. We've heard of cases where
initial setup costs were charged. If not much manual intervention is needed a
\TEX\ solution mostly concerns initial costs.

Let's return to the main question \quotation {Why use \TEX ?} in which you can
replace \TEX\ by one of the macro packages build on top of it, for instance
\CONTEXT. If an (somewhat older) organization considers using \TEX\ it should
also ask itself, why it wasn't considered long ago already? For sure there have
been developments in \TEX\ engines (in \CONTEXT\ we use \LUATEX) as well as
possibilities of macro packages but if you look at the documents produced with
them, there is not that much difference with decades ago. Processing has become
faster, some things have become easier, but new technologies have always been
supported as soon at they showed up. Advertising is often just repeating an old
message.

The \TEX\ ecosystem was among the first in supporting for instance \OPENTYPE, and
the community even made sure that there were free fonts available. A format like
\PDF\ was supported as soon as it shows up and \TEX\ was the first to demonstrate
what advanced features were there and how way it was to adapt to changes.
Processing \XML\ using \TEX\ has never been a big deal and if that is a reason to
look at this already old and mature technology, then an organization can wonder
if years and opportunities (for instance for publishing on demand or easy
updating of manuals) have been lost. Of course there are (and have been)
alternative tools but the arguments for using \TEX\ or not are not much different
now. It can be bad marketing of open and free software. It can be that \TEX\ has
been around too long. It can also be that its message was not understood yet. On
the other hand, in software development it's quite common to reinvent wheels and
present old as new. It's never to late to catch on.

\stopsection

\stopchapter

\stopcomponent
