% language=uk
%
% copyright=pragma-ade readme=readme.pdf licence=cc-by-nc-sa

\startcomponent metafun-layout

\environment metafun-environment

\startchapter[title={Enhancing the layout}]

\startintro

One of the most powerful and flexible commands of \CONTEXT\ is \type {\framed}.
We can use the background features of this command to invoke and position
graphics that adapt themselves to the current situation. Once understood,
overlays will become a natural part of the \CONTEXT\ users toolkit.

\stopintro

\startsection[title={Overlays}]

\index{overlays}

Many \CONTEXT\ commands support overlays. The term {\em overlay} is a bit
confusing, since such an overlay in most cases will lay under the text. However,
because there can be many layers on top of each other, the term suits its
purpose.

When we want to put a \METAPOST\ graphic under some text, we go through a three
step process. First we define the graphic itself:

\startbuffer
\startuniqueMPgraphic{demo circle}
  path p ;
  p := fullcircle xscaled \overlaywidth yscaled \overlayheight ;
  fill p withcolor .85white ;
  draw p withpen pencircle scaled 2pt withcolor .625red ;
\stopuniqueMPgraphic
\stopbuffer

\typebuffer

\getbuffer

This graphic will adapt itself to the width and height of the overlay. Both \type
{\overlaywidth} and \type {\overlayheight} are macros that return a dimension
followed by a space. The next step is to register this graphic as an overlay.

\startbuffer
\defineoverlay[demo circle][\uniqueMPgraphic{demo circle}]
\stopbuffer

\typebuffer

\getbuffer

We can now use this overlay in any command that provides the \type {\framed}
functionality. Since this graphic is defined as unique, \CONTEXT\ will try to
reuse already calculated and embedded graphics when possible.

\startbuffer
\framed[background=demo circle]{This text is overlayed.}
\stopbuffer

\typebuffer

The background can be set to \type {color}, \type {screen}, an overlay
identifier, like \typ {demo circle}, or a comma separated list of those.

\startlinecorrection[blank]
\getbuffer
\stoplinecorrection

The \type {\framed} command automatically draws a ruled box, which can be quite
useful when debugging a graphic. However, in this case we want to turn the frame
off.

\startbuffer
\framed
  [background=demo circle,frame=off]
  {This text is overlayed.}
\stopbuffer

\typebuffer

\startlinecorrection[blank]
\getbuffer
\stoplinecorrection

In this case, it would have made sense to either set the \type {offset} to a
larger value, or to set \type {backgroundoffset}. In the latter case, the ellipse
is positioned outside the frame.

The difference between the three offsets \type {offset}, \type {frameoffset} and
\type {backgroundoffset} is demonstrated in \in {figure} [fig:offsets]. While the
\type {offset} is added to the (natural or specified) dimensions of the content
of the box, the other two are applied to the frame and background and don't add
to the dimensions.

In the first row we only set the \type {offset}, while in the second row, the
(text) offset is set to \type {3pt}. When not specified, the \type {offset} has a
comfortable default value of \type {.25ex} (some 25\% of the height of an~x).

\startbuffer
\setupframed
  [width=.3\textwidth,
   background=demo circle]
\startcombination[3*3]
  {\framed[offset=none]         {\TeX}} {\tt offset=none}
  {\framed[offset=overlay]      {\TeX}} {\tt offset=overlay}
  {\framed[offset=0pt]          {\TeX}} {\tt offset=0pt}
  {\framed[offset=1pt]          {\TeX}} {\tt offset=1pt}
  {\framed[offset=2pt]          {\TeX}} {\tt offset=2pt}
  {\framed[offset=4pt]          {\TeX}} {\tt offset=4pt}
  {\framed[offset=3pt]          {\TeX}} {\tt offset=3pt}
  {\framed[frameoffset=3pt]     {\TeX}} {\tt frameoffset=3pt}
  {\framed[backgroundoffset=3pt]{\TeX}} {\tt backgroundoffset=3pt}
\stopcombination
\stopbuffer

\typebuffer

\placefigure
  [here][fig:offsets]
  {The three offsets.}
  {\getbuffer}

As the first row in \in {figure} [fig:offsets] demonstrates, instead of a value,
one can pass a keyword. The \type {overlay} keyword implies that there is no
offset at all and that the lines cover the content. With \type {none} the frame
is drawn tight around the content. When the offset is set to \type {0pt} or more,
the text is automatically set to at least the height of a line. You can turn this
feature off by saying \type {strut=off}. More details can be found in the
\CONTEXT\ manual.

In \in {figure} [fig:all offsets] we have set {offset} to \type {3pt}, \type
{frameoffset} to \type {6pt} and \type {backgroundoffset} to \type {9pt}. Both
the frame and background offset are sort of imaginary, since they don't
contribute to the size of the box.

\startbuffer
\ruledhbox
  {\framed
     [offset=3pt,frameoffset=6pt,backgroundoffset=9pt,
      background=screen,backgroundscreen=.85]
     {Welcome in the hall of frame!}}
\stopbuffer

\typebuffer

\placefigure
  [here][fig:all offsets]
  {The three offsets.}
  {\getbuffer}

\stopsection

\startsection[title={Overlay variables}]

The communication between \TEX\ and embedded \METAPOST\ graphics takes place by
means of some macros.

\starttabulate[|l|p|]
\HL
\NC overlay status macro    \NC meaning                        \NC \NR
\HL
\NC \tex {overlaywidth}     \NC the width of the graphic, as
                                calculated from the actual
                                width and background offset    \NC \NR
\NC \tex {overlayheight}    \NC the height of the graphic, as
                                calculated from the actual
                                height, depth and background
                                offset                         \NC \NR
\NC \tex {overlaydepth}     \NC the depth of the graphic, if
                                available                      \NC \NR
\NC \tex {overlaycolor}     \NC the background color, if given \NC \NR
\NC \tex {overlaylinecolor} \NC the color of the frame         \NC \NR
\NC \tex {overlaylinewidth} \NC the width of the frame         \NC \NR
\HL
\stoptabulate

The dimensions of the overlay are determined by dimensions of the background,
which normally is the natural size of a \type {\framed}. When a background offset
is specified, it is added to \type {overlayheight} and \type {overlaywidth}.

Colors can be converted by \type {\MPcolor} and in addition to the macros
mentioned, you can use all macros that expand into a dimension or dimen register
prefixed by the \TEX\ primitive \type {\the} (this and other primitives are
explained in \quotation {The \TeX book}, by Donald Knuth).

\stopsection

\startsection[title={Stacking overlays}]

A background can be a gray scale (\type {screen}), a color (\type {color}), a
previously defined overlay identifier, or any combination of these. The next
assignments are therefore valid:

\starttyping
\framed[background=color,backgroundcolor=red]{...}
\framed[background=screen,backgroundscreen=.8]{...}
\framed[background=circle]{...}
\framed[background={color,cow},backgroundcolor=red]{...}
\framed[background={color,cow,grid},backgroundcolor=red]{...}
\stoptyping

In the last three cases of course you have to define \type {circle}, \type {cow}
and \type {grid} as overlay. These items are packed in a comma separated list,
which has to be surrounded by \type {{}}.

\stopsection

\startsection[title={Foregrounds}]

\startbuffer[a]
\startuniqueMPgraphic{backfore}
  draw fullcircle
    xscaled \overlaywidth yscaled \overlayheight
    withpen pencircle scaled 2pt
    withcolor .625yellow ;
\stopuniqueMPgraphic

\defineoverlay[backfore][\uniqueMPgraphic{backfore}]
\stopbuffer

\startbuffer[b]
\framed
  [background=backfore,backgroundoffset=4pt]
  {one, two, three, \unknown}
\stopbuffer

\startbuffer[c]
\framed
  [background={foreground,backfore},backgroundoffset=4pt]
  {one, two, three, \unknown}
\stopbuffer

The overlay system is actually a system of layers. Sometimes we are confronted
with a situation in which we want the text behind another layer. This can be
achieved by explicitly placing the foreground layer, as in \in {figure}
[fig:foreground].

\getbuffer[a]

\placefigure
  [here][fig:foreground]
  {Foreground material moved backwards.}
  {\setupframed[linewidth=1pt]%
   \startcombination
     {\getbuffer[b]} {frame on top layer}
     {\getbuffer[c]} {frame on bottom layer}
   \stopcombination}

The graphic layer is defined as follows:

\typebuffer[a]

The two framed texts have a slightly different definition. The leftmost graphic
is defined as:

\typebuffer[b]

The rightmost graphic is specified as:

\typebuffer[c]

The current values of the frame color and frame width are passed to the overlay.
It often makes more sense to use colors defined at the document level, if only to
force consistency.

\startbuffer
\startuniqueMPgraphic{super ellipse}
  path p ; p := unitsquare
    xscaled \overlaywidth yscaled \overlayheight
    superellipsed .85 ;
  pickup pencircle scaled \overlaylinewidth ;
  fill p withcolor \MPcolor{\overlaycolor} ;
  draw p withcolor \MPcolor{\overlaylinecolor} ;
\stopuniqueMPgraphic

\defineoverlay[super ellipse][\uniqueMPgraphic{super ellipse}]
\stopbuffer

\typebuffer \getbuffer

This background demonstrates that a super ellipse is rather well suited as frame.

\startbuffer
\framed
  [background=super ellipse,
   frame=off,
   width=3cm,
   align=middle,
   framecolor=darkyellow,
   rulethickness=2pt,
   backgroundcolor=darkgray]
  {\white This is a\\Super Ellipsed\\sentence.}
\stopbuffer

\typebuffer

Such a super ellipse looks quite nice and is a good candidate for backgrounds,
for which the superness should be at least~.85.

\startlinecorrection[blank]
\getbuffer
\stoplinecorrection

\stopsection

\startsection[title={Typesetting graphics}]

I have run into people who consider it kind of strange when you want to use \TEX\
for non||mathematical typesetting. If you agree with them, you may skip this
section with your eyes closed.

One of the \CONTEXT\ presentation styles (number 15, tagged as balls) stepwise
builds screens full of sentences, quotes or concepts, packages in balloons and
typesets them as a paragraph. We will demonstrate that \TEX\ can typeset graphics
using the following statement.

% \let\processword\relax

\startbuffer[lions]
\processwords{As you may know, \TEX's ambassador is a lion, while {\METAFONT}
is represented by a lioness. It is still unclear if they have a relationship,
but if so, and if a cub is born, may it enjoy \METAFUN.}
\stopbuffer

\startquotation
\def\processwords#1{#1}\getbuffer[lions]
\stopquotation

The low level \CONTEXT\ macro \type {\processwords} provides a mechanism to treat
the individual words of its argument. The macro is called as follows:

\typebuffer[lions]

In order to perform a task, you should also define a macro \type {\processword},
which takes one argument. The previous quote was typeset with the following
definition in place:

\starttyping
\def\processword#1{#1}
\stoptyping

A slightly more complicated definition is the following:

\startbuffer
\def\processword#1{\noindent\framed{#1}\space}
\stopbuffer

\typebuffer \getbuffer

We now get:

\blank\getbuffer[lions]\blank

If we can use \type {\framed}, we can also use backgrounds.

\startbuffer
\def\processword#1%
  {\noindent\framed[frame=off,background=lions]{#1} }
\stopbuffer

\typebuffer \getbuffer

We can add a supperellipsed frame using the following definition:

\startbuffer
\startuniqueMPgraphic{lions a}
  path p ; p := fullsquare
   xyscaled (\overlaywidth,\overlayheight) superellipsed .85 ;
  pickup pencircle scaled 1pt ;
  fill p withcolor .850white ; draw p withcolor .625yellow ;
\stopuniqueMPgraphic

\defineoverlay[lions][\uniqueMPgraphic{lions a}]
\stopbuffer

\typebuffer \getbuffer

\bgroup \blank \veryraggedright\getbuffer[lions]\unskip \blank \egroup

\startbuffer
\startuseMPgraphic{lions b}
  path p ; p := fullsquare
    xyscaled (\overlaywidth,\overlayheight) randomized 5pt ;
  pickup pencircle scaled 1pt ;
  fill p withcolor .850white ; draw p withcolor .625yellow ;
\stopuseMPgraphic

\defineoverlay[lions][\uniqueMPgraphic{lions b}]
\stopbuffer

\typebuffer \getbuffer

\bgroup \blank \veryraggedcenter\getbuffer[lions]\unskip \blank \egroup

\startbuffer
\startuniqueMPgraphic{lions c}
  path p ; p := fullsquare
    xyscaled (\overlaywidth,\overlayheight) squeezed 2pt ;
  pickup pencircle scaled 1pt ;
  fill p withcolor .850white ; draw p withcolor .625yellow ;
\stopuniqueMPgraphic

\defineoverlay[lions][\uniqueMPgraphic{lions c}]
\stopbuffer

\typebuffer \getbuffer

\bgroup \blank \veryraggedleft\getbuffer[lions]\unskip \blank \egroup

These paragraphs were typeset with the following settings.

\starttyping
\setupalign[broad, right]  % == \veryraggedright
\setupalign[broad, middle] % == \veryraggedcenter
\setupalign[broad, left]   % == \veryraggedleft
\stoptyping

The \type {broad} increases the raggedness. We defined three different graphics
(a, b and c) because we want some to be unique, which saves some processing. Of
course we don't reuse the random graphics. In the definition of \type
{\processword} we have to use \type {\noindent} because otherwise \TEX\ will put
each graphic on a line of its own. Watch the space at the end of the macro.

\stopsection

\startsection[title={Graphics and macros}]

Because \TEX's typographic engine and \METAPOST's graphic engine are separated,
interfacing between them is not as natural as you may expect. In \CONTEXT\ we
have tried to integrate them as much as possible, but using the interface is not
always as convenient as it should be. What method you follow, depends on the
problem at hand.

The official \METAPOST\ way to embed \TEX\ code into graphics is to use \typ
{btex ... etex}. As soon as \CONTEXT\ writes the graphic data to the intermediate
\METAPOST\ file, it looks for these commands. When it has encountered an \type
{etex}, \CONTEXT\ will make sure that the text that is to be typeset by \TEX\ is
{\em not} expanded. This is what you may expect, because when you would embed
those commands in a stand||alone graphic, they would also not be expanded, if
only because \METAPOST\ does not know \TEX. With expansion we mean that \TEX\
commands are replaced by their meaning (which can be quite extensive).

{\em Users of \CONTEXT\ MKIV\ can skip the next paragraph}.

When \METAPOST\ sees a \type {btex} command, it will consult a so called \type
{mpx} file. This file holds the \METAPOST\ representation of the text typeset by
\TEX. Before \METAPOST\ processes a graphic definition file, it first calls
another program that filters the \type {btex} commands from the source file, and
generates a \TEX\ file from them. This file is then processed by \TEX, and after
that converted to a \type {mpx} file. In \CONTEXT\ we let \TEXEXEC\ take care of
this whole process.

Because the \typ {btex ... etex} commands are filtered from the raw \METAPOST\
source code, they cannot be part of macro definitions and loop constructs. When
used that way, only one instance would be found, while in practice multiple
instances may occur.

This drawback is overcome by \METAFUN's \type {textext} command. This command
still uses \typ {btex ... etex} but writes these commands to a separate job
related file each time it is used. \footnote {It took the author a while to find
out that there is a \METAPOST\ module called \type {tex.mp} that provides a
similar feature, but with the disadvantage that each text results in a call to
\TEX. Each text goes into a temporary file, which is then included and results in
\METAPOST\ triggering \TEX.} After the first \METAPOST\ run, this file is merged
with the original file, and \METAPOST\ is called again. So, at the cost of an
additional run, we can use text typeset by \TEX\ in a more versatile way. Because
\METAPOST\ runs are much faster than \TEX\ runs, the price to pay in terms of run
time is acceptable. Unlike \typ {btex ... etex}, the \TEX\ code in \type
{textext} command is expanded, but as long as \CONTEXT\ is used this is seldom a
problem, because most commands are somewhat protected.

If we define a graphic with text to be typeset by \TEX, there is a good chance
that this text is not frozen but passes as argument. A \TEX||like solution for
passing arbitrary content to such a graphic is the following: \footnote {The \type
{\unexpanded} prefix makes the command robust for being passed as argument. It is not
to be confused with the primitive. We had this feature already when the primitive
showed up and it was considered to be inconvenient for other macro packages to adapt to
the \CONTEXT\ situation. So keep that in mind when you mix macro packages.}

\startbuffer[def]
\unexpanded\def\RotatedText#1#2%
  {\startuseMPgraphic{RotatedText}
     draw btex #2 etex rotated #1 ;
   \stopuseMPgraphic
   \useMPgraphic{RotatedText}}
\stopbuffer

\typebuffer[def] \getbuffer[def]

This macro takes two arguments (the \type {#} identifies an
argument):

\startbuffer[exa]
\RotatedText{15}{Some Rotated Text}
\stopbuffer

\typebuffer[exa]

The text is rotated over 15 degrees about the origin in a counterclockwise
direction.

\startlinecorrection[blank]
\getbuffer[exa]
\stoplinecorrection

In \CONTEXT\ we seldom pass settings like the angle of rotation in this manner.
You can use \type {\setupMPvariables} to set up graphic||specific variables. Such
a variable can be accessed with \type {\MPvar}.

\startbuffer[def]
\setupMPvariables[RotatedText][rotation=90]

\startuseMPgraphic{RotatedText}
  draw textext{Some Text} rotated \MPvar{rotation} ;
\stopuseMPgraphic
\stopbuffer

\typebuffer[def] \getbuffer[def]

An example:

\startbuffer[exa]
\RotatedText{-15}{Some Rotated Text}
\stopbuffer

\typebuffer[exa]

\startlinecorrection[blank]
\getbuffer[exa]
\stoplinecorrection

In a similar fashion we can isolate the text. This permits us to use the same
graphics with different settings.

\startbuffer[def]
\setupMPvariables[RotatedText][rotation=270]

\setMPtext{RotatedText}{Some Text}

\startuseMPgraphic{RotatedText}
  draw \MPbetex{RotatedText} rotated \MPvar{rotation} ;
\stopuseMPgraphic
\stopbuffer

\typebuffer[def] \getbuffer[def]

This works as expected:

\startbuffer[exa]
\RotatedText{165}{Some Rotated Text}
\stopbuffer

\typebuffer[exa]

\startlinecorrection[blank]
\getbuffer[exa]
\stoplinecorrection

It is now a small step towards an encapsulating macro (we assume that you are
familiar with \TEX\ macro definitions).

\startbuffer[def]
\def\RotatedText[#1]#2%
  {\setupMPvariables[RotatedText][#1]%
   \setMPtext{RotatedText}{#2}%
   \useMPgraphic{RotatedText}}

\setupMPvariables[RotatedText][rotation=90]

\startuseMPgraphic{RotatedText}
  draw \MPbetex{RotatedText} rotated \MPvar{rotation} ;
\stopuseMPgraphic
\stopbuffer

\typebuffer[def] \getbuffer[def]

Again, we default to a 90 degrees rotation, and pass both the settings and text
in an indirect way. This method permits you to build complicated graphics and
still keep macros readable.

\startbuffer[exa]
\RotatedText[rotation=240]{Some Rotated Text}
\stopbuffer

\typebuffer[exa]

\startlinecorrection[blank]
\getbuffer[exa]
\stoplinecorrection

You may wonder why we don't use the variable mechanism to pass the text. The main
reason is that the text mechanism offers a few more features, one of which is
that it passes the text straight on, without the danger of unwanted expansion of
embedded macros. Using \type {\setMPtext} also permits you to separate \TEX\ and
\METAPOST\ code and reuse it multiple times (imagine using the same graphic in a
section head command).

There are three ways to access a text defined with \type {\setMPtext}. Imagine
that we have the following definitions:

\startbuffer
\setMPtext {1} {Now is this \TeX\ or not?}
\setMPtext {2} {See what happens here.}
\setMPtext {3} {Text streams become pictures.}
\stopbuffer

\typebuffer \getbuffer

The \type {\MPbetex} macro returns a \typ {btex ... etex} construct. The \type
{\MPstring} returns the text as a \METAPOST\ string, between quotes. The raw text
can be fetched with \type {\MPtext}.

\startbuffer
\startMPcode
  picture p ; p :=           \MPbetex {1}          ;
  picture q ; q :=  textext( \MPstring{2}        ) ;
  picture r ; r := thelabel("\MPtext  {3}",origin) ;

  for i=p, boundingbox p : draw i withcolor .625red    ; endfor ;
  for i=q, boundingbox q : draw i withcolor .625yellow ; endfor ;
  for i=r, boundingbox r : draw i withcolor .625white  ; endfor ;

  currentpicture := currentpicture scaled 2 ;
  draw origin
    withpen pencircle scaled 5.0mm withcolor white ;
  draw origin
    withpen pencircle scaled 2.5mm withcolor black ;
  draw boundingbox currentpicture
    withpen pencircle scaled .1mm
    dashed evenly ;
\stopMPcode
\stopbuffer

\typebuffer

The first two lines return text typeset by \TEX, while the last line leaves this
to \METAPOST.

\startlinecorrection[blank]
\getbuffer
\stoplinecorrection

If you watch closely, you will notice that the first (red) alternative is
positioned with the baseline on the origin.

\startbuffer
\startMPcode
  picture p ; p :=                  \MPbetex {1}          ;
  picture q ; q :=  textext.origin( \MPstring{2}        ) ;
  picture r ; r := thelabel.origin("\MPtext  {3}",origin) ;

  for i=p, boundingbox p : draw i withcolor .625red    ; endfor ;
  for i=q, boundingbox q : draw i withcolor .625yellow ; endfor ;
  for i=r, boundingbox r : draw i withcolor .625white  ; endfor ;

  currentpicture := currentpicture scaled 2 ;
  draw origin withpen pencircle scaled 5.0mm
    withcolor white ;
  draw origin withpen pencircle scaled 2.5mm
    withcolor black ;
  draw boundingbox currentpicture
    withpen pencircle scaled .1mm
    dashed evenly ;
\stopMPcode
\stopbuffer

\typebuffer

This draws:

\startlinecorrection[blank]
\getbuffer
\stoplinecorrection

This picture demonstrates that we can also position \type {textext}'s and \type
{label}'s on the baseline. For this purpose the repertoire of positioning
directives (\type {top}, \type {lft}, etc.) is extended with an \type {origin}
directive. Another extra positioning directive is \type {raw}. This one does not
do any positioning at all.

\starttyping
picture q ; q :=  textext.origin( \MPstring{2}        ) ;
picture r ; r := thelabel.origin("\MPtext  {3}",origin) ;
\stoptyping

We will now apply this knowledge of text inclusion in graphics to a more advanced
example. The next definitions are the answer to a question on the \CONTEXT\
mailinglist with regards to framed texts with titles.

\startbuffer[a]
\defineoverlay[FunnyFrame][\useMPgraphic{FunnyFrame}]

\defineframedtext[FunnyText][frame=off,background=FunnyFrame]

\def\StartFrame{\startFunnyText}
\def\StopFrame {\stopFunnyText }

\def\FrameTitle#1%
  {\setMPtext{FunnyFrame}{\hbox spread 1em{\hss\strut#1\hss}}}

\setMPtext{FunnyFrame}{} % initialize the text variable
\stopbuffer

\startbuffer[b]
\startuseMPgraphic{FunnyFrame}
  picture p ; numeric w, h, o ;
  p := textext.rt(\MPstring{FunnyFrame}) ;
  w := OverlayWidth ; h := OverlayHeight ; o := BodyFontSize ;
  p := p shifted (2o,h-ypart center p) ; draw p ;
  drawoptions (withpen pencircle scaled 1pt withcolor .625red) ;
  draw (2o,h)--(0,h)--(0,0)--(w,0)--(w,h)--(xpart urcorner p,h) ;
  draw boundingbox p ;
  setbounds currentpicture to unitsquare xyscaled(w,h) ;
\stopuseMPgraphic
\stopbuffer

\startbuffer[c1]
\FrameTitle{Zapf (1)}

\StartFrame
Coming back to the use of typefaces in electronic
publishing: many of the new typographers receive their
knowledge and information about the rules of typography from
books, from computer magazines or the instruction manuals
which they get with the purchase of a PC or software.
\StopFrame
\stopbuffer

\getbuffer[a,b,c1]

In this example, the title is positioned on top of the frame. Title and text are
entered as:

\typebuffer[c1]

The implementation is not that complicated and uses the frame commands that are
built in \CONTEXT. Instead of letting \TEX\ draw the frame, we use \METAPOST,
which we also use for handling the title. The graphic is defined as follows:

\typebuffer[b]

Because the framed title is partly positioned outside the main frame, and because
the main frame will be combined with the text, we need to set the boundingbox
explicitly. This is a way to create so called free figures, where part of the
figure lays beyond the area that is taken into account when positioning the
graphic. The shift

\starttyping
... shifted (2o,h-ypart center p)
\stoptyping

ensures that the title is vertically centered over the top line of the main box.

The macros that use this graphic combine some techniques of defining macros,
using predefined \CONTEXT\ classes, and passing information to graphics.

\typebuffer[a]

There is a little bit of low level \TEX\ code involved, like a horizontal box
(\type {\hbox}) that stretches one em||space beyond its natural size (\type
{spread 1em}) with a centered text (two times \type {\hss}). Instead of applying
this spread, we could have enlarged the frame on both sides.

\startbuffer[b]
\startuseMPgraphic{FunnyFrame}
  picture p ; numeric o ; path a, b ; pair c ;
  p := textext.rt(\MPstring{FunnyFrame}) ;
  a := unitsquare xyscaled(OverlayWidth,OverlayHeight) ;
  o := BodyFontSize ;
  p := p shifted (2o,OverlayHeight-ypart center p) ;
  drawoptions (withpen pencircle scaled 1pt withcolor .625red) ;
  b := a randomized (o/2) ;
  fill b withcolor .85white ; draw b ;
  b := (boundingbox p) randomized (o/8) ;
  fill b withcolor .85white ; draw b ;
  draw p withcolor black;
  setbounds currentpicture to a ;
 \stopuseMPgraphic
\stopbuffer

In the previous graphic we calculated the big rectangle taking the small one into
account. This was needed because we don't use a background fill. The next
definition does, so there we can use a more straightforward approach by just
drawing (and filling) the small rectangle on top of the big one.

\typebuffer[b] \getbuffer[b]

\startbuffer[c2]
\FrameTitle{Zapf (2)}

\StartFrame
There is not so much basic instruction, as of now, as there
was in the old days, showing the differences between good
and bad typographic design.
\StopFrame
\stopbuffer

\getbuffer[c2]

Because we use a random graphic, we cannot guarantee beforehand that the left and
right edges of the small shape touch the horizontal lines in a nice way. The next
alternative displaces the small shape plus text so that its center lays on the
line. On the average, this looks better.

\startbuffer[b]
\startuseMPgraphic{FunnyFrame}
  picture p ; numeric o ; path a, b ; pair c ;
  p := textext.rt(\MPstring{FunnyFrame}) ;
  a := unitsquare xyscaled(OverlayWidth,OverlayHeight) ;
  o := BodyFontSize ;
  p := p shifted (2o,OverlayHeight-ypart center p) ;
  drawoptions (withpen pencircle scaled 1pt withcolor .625red) ;
  b := a randomized (o/2) ;
  fill b withcolor .85white ; draw b ;
  c := center p ;
  c := b intersectionpoint (c shifted (0,-o)--c shifted(0,o)) ;
  p := p shifted (c-center p) ;
  b := (boundingbox p) randomized (o/8) ;
  fill b withcolor .85white ; draw b ;
  draw p withcolor black;
  setbounds currentpicture to a ;
 \stopuseMPgraphic
\stopbuffer

\typebuffer[b] \getbuffer[b]

\getbuffer[c2]

Yet another definition uses super ellipsed shapes instead of random ones. We need
a high degree of superness (.95) in order to make sure that the curves don't
touch the texts.

\startbuffer[b]
\startuseMPgraphic{FunnyFrame}
  picture p ; numeric o ; path a, b ; pair c ;
  p := textext.rt(\MPstring{FunnyFrame}) ;
  o := BodyFontSize ;
  a := unitsquare xyscaled(OverlayWidth,OverlayHeight) ;
  p := p shifted (2o,OverlayHeight-ypart center p) ;
  drawoptions (withpen pencircle scaled 1pt withcolor .625red) ;
  b := a superellipsed .95 ;
  fill b withcolor .85white ; draw b ;
  b := (boundingbox p) superellipsed .95 ;
  fill b withcolor .85white ; draw b ;
  draw p withcolor black ;
  setbounds currentpicture to a ;
\stopuseMPgraphic
\stopbuffer

\typebuffer[b] \getbuffer[b]

\startbuffer[c3]
\FrameTitle{Zapf (3)}

\StartFrame
Many people are just fascinated by their PC's tricks, and
think that a widely||praised program, called up on the
screen, will make everything automatic from now on.
\StopFrame
\stopbuffer

\getbuffer[c3]

There are quite some hard coded values in these graphics, like the linewidths,
offsets and colors. Some of these can be fetched from the \type {\framed}
environment either by using \TEX\ macros or dimensions, or by using their
\METAFUN\ counterparts. In the following table we summarize both the available
\METAPOST\ variables and their \TEX\ counterparts. They may be used
interchangeably.

\starttabulate[|l|Tl|l|]
\HL
\NC \bf \METAPOST\ variable \NC \rm \bf \TEX\ command \NC \bf meaning \NC\NR
\HL
\NC OverlayWidth     \NC \string\overlaywidth
                     \NC current width \NC\NR
\NC OverlayHeight    \NC \string\overlayheight
                     \NC current height \NC\NR
\NC OverlayDepth     \NC \string\overlayheight
                     \NC current depth (often zero) \NC\NR
\NC OverlayColor     \NC \string\MPcolor\string{\string\overlaycolor\string}
                     \NC background color \NC\NR
\NC OverlayLineWidth \NC \string\overlaylinewidth
                     \NC width of the frame \NC\NR
\NC OverlayLineColor \NC \string\MPcolor\string{\overlaylinecolor\string}
                     \NC color of the frame \NC\NR
\NC BaseLineSkip     \NC \string\the\string\baselineskip
                     \NC main line distance \NC\NR
\NC LineHeight       \NC \string\the\string\baselineskip
                     \NC idem \NC\NR
\NC BodyFontSize     \NC \string\the\string\bodyfontsize
                     \NC font size of the running text \NC\NR
\NC StrutHeight      \NC \string\strutheight
                     \NC space above the baseline \NC\NR
\NC StrutDepth       \NC \string\strutdepth
                     \NC space below the baseline \NC\NR
\NC ExHeight         \NC 1ex
                     \NC height of an x \NC \NR
\NC EmWidth          \NC 1em
                     \NC width of an m-dash \NC \NR
\HL
\stoptabulate

\startbuffer[b]
\startuseMPgraphic{FunnyFrame}
  picture p ; numeric o ; path a, b ; pair c ;
  p := textext.rt(\MPstring{FunnyFrame}) ;
  o := BodyFontSize ;
  a := unitsquare xyscaled(OverlayWidth,OverlayHeight) ;
  p := p shifted (2o,OverlayHeight-ypart center p) ;
  pickup pencircle scaled OverlayLineWidth ;
  b := a superellipsed .95 ;
  fill b withcolor OverlayColor ;
  draw b withcolor OverlayLineColor ;
  b := (boundingbox p) superellipsed .95 ;
  fill b withcolor OverlayColor ;
  draw b withcolor OverlayLineColor ;
  draw p withcolor black ;
  setbounds currentpicture to a ;
\stopuseMPgraphic
\stopbuffer

\typebuffer[b] \getbuffer[b]

\startbuffer[d]
\setupframedtexts
  [FunnyText]
  [backgroundcolor=lightgray,
   framecolor=darkred,
   rulethickness=2pt,
   offset=\bodyfontsize,
   before={\blank[big,medium]},
   after={\blank[big]},
   width=\textwidth]
\stopbuffer

\getbuffer[d,c3]

We used the following command to pass the settings:

\typebuffer[d]

In a real implementation, we should also take care of some additional spacing
before the text, which is why we have added more space before than after the
framed text.

We demonstrated that when defining graphics that are part of the layout, you need
to have access to information known to the typesetting engine. Take \in {figure}
[fig:penalty]. The line height needs to match the font and the two thin
horizontal lines should match the {\em x}||height. We also need to position the
baseline, being the lowest one of a pair of lines, in such a way that it suits
the proportions of the line as specified by the strut. A strut is an imaginary
large character with no width. You should be aware of the fact that while \TEX\
works its way top||down, in \METAPOST\ the origin is in the lower left corner.

\startmode[screen]

\placefigure
  [page][fig:penalty]
  {Penalty lines.}
  {\typesetfile[mfun-902.tex][page=1,frame=on,height=.85\textheight]}

\stopmode

\startnotmode[screen]

\placefigure
  [here][fig:penalty]
  {Penalty lines.}
  {\typesetfile[mfun-902.tex][page=1,frame=on,height=.50\textheight]}

\stopnotmode

\typebuffer[handwrit]

This code demonstrates the use of \type {LineHeight}, \type {ExHeight}, \type
{StrutHeight} and \type {StrutDepth}. We set the interline spacing to 1.5 so that
we get a bit more loose layout. The variables mentioned are set each time a
graphic is processed and thereby match the current font settings.

\stopsection

\stopchapter

\stopcomponent
