% language=us

% \hfil \hss
% spread

\environment lowlevel-style

\startdocument
  [title=boxes,
   color=middlered]

\startsection[title=Preamble]

\startsubsection[title=Introduction]

An average \CONTEXT\ user will not use the low level box primitives but a basic
understanding of how \TEX\ works doesn't hurt. In fact, occasionally using a box
command might bring a solution not easily achieved otherwise, simply because a
more high level interface can also be in the way.

The best reference is of course The \TeX book so if you're really interested in
the details you should get a copy of that book. Below I will not go into details
about all kind of glues, kerns and penalties, just boxes it is.

This explanation will be extended when I feel the need (or users have questions
that can be answered here).

\stopsubsection

\startsubsection[title=Boxes]

This paragraph of text is made from lines that contain words that themselves
contain symbolic representations of characters. Each line is wrapped in a so
called horizontal box and eventually those lines themselves get wrapped in what
we call a vertical box.

\startbuffer
\vbox \bgroup
    \hsize 5cm
    \raggedright
    This is a rather narrow paragraph blown up a bit. Here we use a flush left,
    aka ragged right, approach.
\egroup
\stopbuffer

When we expose some details of a paragraph it looks like this:

\startlinecorrection
\startcombination[2*1]
    {\scale[width=8cm]{\showmakeup[boxes]\getbuffer}} {}
    {\scale[width=8cm]{\showmakeup\getbuffer}}        {}
\stopcombination
\stoplinecorrection

The left only shows the boxes, the variant at the right shows (font) kerns and
glue too. Because we flush left, there is rather strong right skip glue at the
right boundary of the box. If font kerns show up depends on the font, not all
fonts have them (or have only a few). The glyphs themselves are also kind of
boxed, as their dimensions determine the area that they occupy:

\startlinecorrection
    \scale[width=\textwidth]{\showglyphs\hbox{This is a rather ...}}
\stoplinecorrection

But, internally they are not really boxed, as they already are a single quantity.
The same is true for rules: they are just blobs with dimensions. A box on the
other hand wraps a linked list of so called nodes: glyphs, kerns, glue,
penalties, rules, boxes, etc. It is a container with properties like width,
height, depth and shift.

\stopsubsection

\stopsection

\startsection[title={\TEX\ primitives}]

The box model is reflected in \TEX's user interface but not by that many
commands, most noticeably \type {\hbox}, \type {\vbox} and \type {\vtop}. Here is
an example of the first one:

\starttyping[option=TEX]
\hbox width 10cm{text}
\hbox width 10cm height 1cm depth 5mm{text}
text \raise5mm\hbox{text} text
\stoptyping

The \type {\raise} and \type {\lower} commands behave the same but in opposite
directions. One could as well have been defined in terms of the other.

\startbuffer
text \raise  5mm \hbox to 2cm {text}
text \lower -5mm \hbox to 2cm {text}
text \raise -5mm \hbox to 2cm {text}
text \lower  5mm \hbox to 2cm {text}
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
{\dontcomplain\showboxes\getbuffer}
\stoplinecorrection

A box can be moved to the left or right but, believe it or not, in \CONTEXT\ we
never use that feature, probably because the consequences for the width are such
that we can as well use kerns. Here are some examples:

\startbuffer
text \vbox{\moveleft  5mm \hbox {left}}text !
text \vbox{\moveright 5mm \hbox{right}}text !
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
{\dontcomplain\getbuffer}
\stoplinecorrection

\startbuffer
text \vbox{\moveleft  25mm \hbox {left}}text !
text \vbox{\moveright 25mm \hbox{right}}text !
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
{\dontcomplain\getbuffer}
\stoplinecorrection

Code like this will produce a complaint about an underfull box but we can easily
get around that:

\startbuffer
text \raise  5mm \hbox to 2cm {\hss text}
text \lower -5mm \hbox to 2cm {text\hss}
text \raise -5mm \hbox to 2cm {\hss text}
text \lower  5mm \hbox to 2cm {text\hss}
\stopbuffer

\typebuffer[option=TEX]

The \type {\hss} primitive injects a glue that when needed will fill up the
available space. So, here we force the text to the right or left.

\startlinecorrection
{\dontcomplain\showboxes\getbuffer}
\stoplinecorrection

We have three kind of boxes: \type {\hbox}, \type {\vbox} and \type {\vtop}:

\startbuffer
\hbox{\strut height and depth\strut}
\vbox{\hsize 4cm \strut height and depth\par and width\strut}
\vtop{\hsize 4cm \strut height and depth\par and width\strut}
\stopbuffer

\typebuffer[option=TEX]

A \type {\vbox} aligns at the bottom and a \type {\vtop} at the top. I have added
some so called struts to enforce a consistent height and depth. A strut is an
invisible quantity (consider it a black box) that enforces consistent line
dimensions: height and depth.


\startlinecorrection
{\dontcomplain\hbox{\showstruts\showboxes\getbuffer}}
\stoplinecorrection

You can store a box in a register but you need to be careful not to use a
predefined one. If you need a lot of boxes you can reserve some for your own:

\starttyping
\newbox\MySpecialBox
\stoptyping

but normally you can do with one of the scratch registers, like 0, 2, 4, 6 or 8,
for local boxes, and 1, 3, 5, 7 and 9 for global ones. Registers are used like:

\starttyping
       \setbox0\hbox{here}
\global\setbox1\hbox{there}
\stoptyping

In \CONTEXT\ you can also use

\starttyping
\setbox\scratchbox   \hbox{here}
\setbox\scratchboxone\hbox{here}
\setbox\scratchboxtwo\hbox{here}
\stoptyping

and some more. In fact, there are quite some predefined scratch registers (boxes,
dimensions, counters, etc). Feel free to investigate further.

When a box is stored, you can consult its dimensions with \type {\wd}, \type
{\ht} and \type {\dp}. You can of course store them for later use.

\starttyping
\scratchwidth \wd\scratchbox
\scratchheight\ht\scratchbox
\scratchdepth \dp\scratchbox
\scratchtotal \dimexpr\ht\scratchbox+\dp\scratchbox\relax
\scratchtotal \htdp\scratchbox
\stoptyping

The last line is \CONTEXT\ specific. You can also set the dimensions

\starttyping
\wd\scratchbox 10cm
\ht\scratchbox 10mm
\dp\scratchbox  5mm
\stoptyping

So you can cheat! A box is placed with \type {\copy}, which keeps the original
intact or \type {\box} which just inserts the box and then wipes the register. In
practice you seldom need a copy, which is more expensive in runtime anyway. Here
we use copy because it serves the examples.

\starttyping
\copy\scratchbox
\box \scratchbox
\stoptyping

\stopsection

\startsection[title={\ETEX\ primitives}]

The \ETEX\ extensions don't add something relevant for boxes, apart from that you
can use the expressions mechanism to mess around with their dimensions. There is
a mechanism for typesetting r2l within a paragraph but that has limited
capabilities and doesn't change much as it's mostly a way to trick the backend
into outputting a stretch of text in the other direction. This feature is not
available in \LUATEX\ because it has an alternative direction mechanism.

\stopsection

\startsection[title={\LUATEX\ primitives}]

The concept of boxes is the same in \LUATEX\ as in its predecessors but there are
some aspects to keep in mind. When a box is typeset this happens in \LUATEX:

\startitemize[n]
    \startitem
        A list of nodes is constructed. In \LUATEX\ this is a double linked
        list (so that it can easily be manipulated in \LUA) but \TEX\ itself
        only uses the forward links.
    \stopitem
    \startitem
        That list is hyphenated, that is: so called discretionary nodes are
        injected. This depends on the language properties of the glyph
        (character) nodes.
    \stopitem
    \startitem
        Then ligatures are constructed, if the font has such combinations. When
        this built|-|in mechanism is used, in \CONTEXT\ we speak of base mode.
    \stopitem
    \startitem
        After that inter|-|character kerns are applied, if the font provides
        them. Again this is a base mode action.
    \stopitem
    \startitem
        Finally the box gets packaged:
        \startitemize
            \startitem
                In the case of a horizontal box, the list is packaged in a
                hlist node, basically one liner, and its dimensions are calculated
                and set.
            \stopitem
            \startitem
                In the case of a vertical box, the paragraph is broken into one
                or more lines, without hyphenation, with optimal hyphenation or
                in the worst case with so called emergency stretch applied, and
                the result becomes a vlist node with its dimensions set.
            \stopitem
        \stopitemize
    \stopitem
\stopitemize

In traditional \TEX\ the first four steps are interwoven but in \LUATEX\ we need
them split because the step~5 can be overloaded by a callback. In that case steps
3 and 4 (and maybe 2) are probably also overloaded, especially when you bring
handling of fonts under \LUA\ control.

New in \LUATEX\ are three packers: \type {\hpack}, \type {\vpack} and \type
{\tpack}, which are companions to \type {\hbox}, \type {\vbox} and \type {\vtop}
but without the callbacks applied. Using them is a bit tricky as you never know
if a callback should be applied, which, because users can often add their own
\LUA\ code, is not something predictable.

Another box related extension is direction. There are four possible directions
but because in \LUAMETATEX\ there are only two. Because this model has been upgraded,
it will be discusses in the next section. A \CONTEXT\ user is supposed to use the
official \CONTEXT\ interfaces in order to be downward compatible.

\stopsection

\startsection[title={\LUAMETATEX\ primitives}]

There are two possible directions: left to right (the default) and right to left
for Hebrew and Arabic. Here is an example that shows how it'd done with low level
directives:

\startbuffer
\hbox direction 0 {from left to right}
\hbox direction 1 {from right to left}
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

A low level direction switch is done with:

\startbuffer
\hbox direction 0
    {from left to right \textdirection 1 from right to left}
\hbox direction 1
    {from right to left \textdirection 1 from left to right}
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

but actually this is kind of {\em not done} in \CONTEXT, because there you are
supposed to use the proper direction switches:

\startbuffer
\naturalhbox {from left to right}
\reversehbox {from right to left}
\naturalhbox {from left to right \righttoleft from right to left}
\reversehbox {from right to left \lefttoright from left to right}
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

Often more is needed to properly support right to left typesetting so using the
\CONTEXT\ commands is more robust.

In \LUAMETATEX\ the box model has been extended a bit, this as a consequence of
dropping the vertical directional typesetting, which never worked well. In
previous sections we discussed the properties width, height and depth and the
shift resulting from a \type {\raise}, \type {\lower}, \type {\moveleft} and
\type {\moveright}. Actually, the shift is also used in for instance positioning
math elements.

The way shifting influences dimensions can be somewhat puzzling. Internally, when
\TEX\ packages content in a box there are two cases:

\startitemize
    \startitem
        When a horizontal box is made, and \typ {height - shift} is larger than the
        maximum height so far, that delta is taken. When \typ {depth + shift} is
        larger than the current depth, then that depth is adapted. So, a shift up
        influences the height and a shift down influences the depth.
    \stopitem
    \startitem
        In the case of vertical packaging, when \typ {width + shift} is larger
        than the maximum box (line) width so far, that maximum gets bumped. So, a
        shift to the right can contribute, but a shift to the left cannot result
        in a negative width. This is also why vertical typesetting, where height
        and depth are swapped with width, goes wrong: we somehow need to map two
        properties onto one and conceptually \TEX\ is really set up for
        horizontal typesetting. (And it's why I decided to just remove it from the
        engine.)
    \stopitem
\stopitemize

This is one of these cases where \TEX\ behaves as expected but it also means that
there is some limitation to what can be manipulated. Setting the shift using one
of the four commands has a direct consequence when a box gets packaged which
happens immediately because the box is an argument to the foursome.

There is in traditional \TEX, probably for good reason, no way to set the shift
of a box, if only because the effect would normally be none. But in \LUATEX\ we
can cheat, and therefore, for educational purposed \CONTEXT\ has implements
some cheats.

We use this sample box:

\startbuffer[demo]
\setbox\scratchbox\hbox\bgroup
    \middlegray\vrule width 20mm depth  -.5mm height 10mm
    \hskip-20mm
    \darkgray  \vrule width 20mm height -.5mm depth   5mm
\egroup
\stopbuffer

\typebuffer[demo][option=TEX]

When we mess with the shift using the \CONTEXT\ \type {\shiftbox} helper, we see
no immediate effect. We only get the shift applied when we use another helper,
\type {\hpackbox}.

\startbuffer
\hbox\bgroup
    \showstruts \strut
    \quad                            \copy\scratchbox
    \quad \shiftbox\scratchbox -20mm \copy\scratchbox
    \quad \hpackbox\scratchbox       \box \scratchbox
    \quad \strut
\egroup
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer[demo]\getbuffer
\stoplinecorrection

When instead we use \type {\vpackbox} we get a different result. This time we
move left.

\startbuffer
\hbox\bgroup
    \showstruts \strut
    \quad                            \copy\scratchbox
    \quad \shiftbox\scratchbox -10mm \copy\scratchbox
    \quad \vpackbox\scratchbox       \copy\scratchbox
    \quad \strut
\egroup
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer[demo]\getbuffer
\stoplinecorrection

The shift is set via \LUA\ and the repackaging is also done in \LUA, using the
low level \type {hpack} and \type {vpack} helpers and these just happen to look
at the shift when doing their job. At the \TEX\ end this never happens.

This long exploration of shifting serves a purpose: it demonstrates that there is
not that much direct control over boxes apart from their three dimensions.
However this was never a real problem as one can just wrap a box in another one
and use kerns to move the embedded box around. But nevertheless I decided to see
if the engine can be a bit more helpful, if only because all that extra wrapping
gives some overhead and complications when we want to manipulate boxes. And of
course it is also a nice playground.

We start with changing the direction. Changing this property doesn't require
repackaging because directions are not really dealt with in the frontend. When
a box is converted to (for instance \PDF) the reversion happens.

\startbuffer
\setbox\scratchbox\hbox{whatever}
\the\boxdirection\scratchbox: \copy\scratchbox \crlf
\boxdirection\scratchbox 1
\the\boxdirection\scratchbox: \copy\scratchbox
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

Another property that can be queried and set is an attribute. In order to get
a private attribute we define one.

\startbuffer
\newattribute\MyAt
\setbox\scratchbox\hbox attr \MyAt 123 {whatever}
[\the\boxattr\scratchbox\MyAt]
\boxattr\scratchbox\MyAt 456
[\the\boxattr\scratchbox\MyAt]
[\ifnum\boxattr\scratchbox\MyAt>400 okay\fi]
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

The sum of the height and depth is available too. Because for practical reasons
setting that property is also needed then, the choice was made to distribute the
value equally over height and depth.

\startbuffer
\setbox\scratchbox\hbox {height and depth}
[\the\ht\scratchbox]
[\the\dp\scratchbox]
[\the\boxtotal\scratchbox]
\boxtotal\scratchbox=20pt
[\the\ht\scratchbox]
[\the\dp\scratchbox]
[\the\boxtotal\scratchbox]
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

We've now arrived to a set of properties that relate to each other. They are
a bit complex and given the number of possibilities one might need to revert
to some trial and error: orientations and offsets. As with the dimensions,
directions and attributes, they are passed as box specification. We start
with the orientation.

\startbuffer
\hbox \bgroup \showboxes
          \hbox orientation 0 {right}
    \quad \hbox orientation 1 {up}
    \quad \hbox orientation 2 {left}
    \quad \hbox orientation 3 {down}
\egroup
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

When the orientation is set, you can also set an offset. Where shifting around a box
can have consequences for the dimensions, an offset is virtual. It gets effective
in the backend, when the contents is converted to some output format.

\startbuffer
\hbox \bgroup \showboxes
          \hbox orientation 0 yoffset  10pt {right}
    \quad \hbox orientation 1 xoffset  10pt {up}
    \quad \hbox orientation 2 yoffset -10pt {left}
    \quad \hbox orientation 3 xoffset -10pt {down}
\egroup
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\getbuffer
\stoplinecorrection

The reason that offsets are related to orientation is that we need to know in
what direction the offsets have to be applied and this binding forces the user to
think about it. You can also set the offsets using commands.

\startbuffer
\setbox\scratchbox\hbox{whatever}%
1                                  \copy\scratchbox
2 \boxorientation\scratchbox 2     \copy\scratchbox
3 \boxxoffset    \scratchbox -15pt \copy\scratchbox
4 \boxyoffset    \scratchbox -15pt \copy\scratchbox
5
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\ruledhbox{\getbuffer}
\stoplinecorrection

\startbuffer
\setbox\scratchboxone\hbox{whatever}%
\setbox\scratchboxtwo\hbox{whatever}%
1 \boxxoffset \scratchboxone -15pt \copy\scratchboxone
2 \boxyoffset \scratchboxone -15pt \copy\scratchboxone
3 \boxxoffset \scratchboxone -15pt \copy\scratchboxone
4 \boxyoffset \scratchboxone -15pt \copy\scratchboxone
5 \boxxmove   \scratchboxtwo -15pt \copy\scratchboxtwo
6 \boxymove   \scratchboxtwo -15pt \copy\scratchboxtwo
7 \boxxmove   \scratchboxtwo -15pt \copy\scratchboxtwo
8 \boxymove   \scratchboxtwo -15pt \copy\scratchboxtwo
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\ruledhbox{\getbuffer}
\stoplinecorrection

The move commands are provides as convenience and contrary to the offsets they do
adapt the dimensions. Internally, with the box, we register the orientation and
the offsets and when you apply these commands multiple times the current values
get overwritten. But \unknown\ because an orientation can be more complex you
might not get the effects you expect when the options we discuss next are used.
The reason is that we store the original dimensions too and these come into play
when these other options are used: anchoring. So, normally you will apply an
orientation and offsets once only.

% the next bit is derived from the followingup document

The orientation specifier is actually a three byte number that best can be seen
hexadecimal (although we stay within the decimal domain). There are three
components: x|-|anchoring, y|-|anchoring and orientation:

\starttyping
0x<X><Y><O>
\stoptyping

or in \TEX\ speak:

\starttyping
"<X><Y><O>
\stoptyping

The landscape and seascape variants both sit on top of the baseline while the
flipped variant has its depth swapped with the height. Although this would be
enough a bit more control is possible.

The vertical options of the horizontal variants anchor on the baseline, lower
corner, upper corner or center.

\startbuffer
\ruledhbox orientation "002 {\TEX} and
\ruledhbox orientation "012 {\TEX} and
\ruledhbox orientation "022 {\TEX} and
\ruledhbox orientation "032 {\TEX}
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\ruledhbox{\getbuffer}
\stoplinecorrection

The horizontal options of the horizontal variants anchor in the center, left,
right, halfway left and halfway right.

\startbuffer
\ruledhbox orientation "002 {\TEX} and
\ruledhbox orientation "102 {\TEX} and
\ruledhbox orientation "202 {\TEX} and
\ruledhbox orientation "302 {\TEX} and
\ruledhbox orientation "402 {\TEX}
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\ruledhbox{\getbuffer}
\stoplinecorrection

The orientation has consequences for the dimensions so they are dealt with in the
expected way in constructing lines, paragraphs and pages, but the anchoring is
virtual, like the offsets. There are two extra variants for orientation zero: on
top of baseline or below, with dimensions taken into account.

\startbuffer
\ruledhbox orientation "000 {\TEX} and
\ruledhbox orientation "004 {\TEX} and
\ruledhbox orientation "005 {\TEX}
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\ruledhbox{\getbuffer}
\stoplinecorrection

The anchoring can look somewhat confusing but you need to keep in mind that it is
normally only used in very controlled circumstances and not in running text.
Wrapped in macros users don't see the details. We're talking boxes here, so for
instance:

\startbuffer
test\quad
\hbox orientation 3 \bgroup
    \strut test\hbox orientation "002 \bgroup\strut test\egroup test%
\egroup \quad
\hbox orientation 3 \bgroup
    \strut test\hbox orientation "002 \bgroup\strut test\egroup test%
\egroup \quad
\hbox orientation 3 \bgroup
    \strut test\hbox orientation "012 \bgroup\strut test\egroup test%
\egroup \quad
\hbox orientation 3 \bgroup
    \strut test\hbox orientation "022 \bgroup\strut test\egroup test%
\egroup \quad
\hbox orientation 3 \bgroup
    \strut test\hbox orientation "032 \bgroup\strut test\egroup test%
\egroup \quad
\hbox orientation 3 \bgroup
    \strut test\hbox orientation "042 \bgroup\strut test\egroup test%
\egroup
\quad test
\stopbuffer

\typebuffer[option=TEX]

\startlinecorrection
\ruledhbox{\getbuffer}
\stoplinecorrection

\stopsection

\stopdocument
